---
title: "Genome-wide Melting Temperature Profiling: An E. coli Case Study"
author: "Junhui Li, Lihua Julie Zhu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genome-wide Melting Temperature Profiling: An E. coli Case Study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 6
)
```

# Introduction

Hasenauer FC, Barreto HC, Lotton C, Matic I (2025). Genome-wide mapping of
spontaneous DNA replication error-hotspots using mismatch repair proteins in
rapidly proliferating *Escherichia coli*. *Nucleic Acids Research*
**53**(2): gkae1196. <doi:10.1093/nar/gkae1196>.

Spontaneous DNA replication errors are not randomly distributed along the
*Escherichia coli* chromosome. Hasenauer *et al.* (2025) mapped these
hotspots genome-wide by tagging the mismatch-repair protein MutL in rapidly
proliferating *mutH*-deficient cells, where mismatches can be detected but
not corrected. MutL ChIP-seq peaks (MutL-associated regions, MutL-AR) are
enriched for sequences of lower thermal stability, mononucleotide repeats
(microsatellites), cruciform-forming DNA, and single-stranded DNA, and are
depleted for GATC methylation sites. These associations motivate a direct
comparison between replication-error hotspots and local DNA melting
temperature (Tm).

This vignette uses **TmCalculator** to compute Tm across the *E. coli* K-12
MG1655 genome (NCBI assembly GCF\_000005845.2 / ASM584v2) for every
non-overlapping 200 bp window -- the bin size used by Hasenauer *et al.*
themselves -- then overlays their
annotation tracks (shipped as `ecoli_rep_hotspots`) to ask whether MutL-AR
windows differ in Tm and GC content from the rest of the chromosome.
Because any window-based Tm profile depends on the chosen bandwidth, we
complement the primary 200 bp analysis with a window-size sensitivity
analysis (50, 100, 200 and 500 bp windows). Briefly, absolute Tm scales
with window length as expected from duplex thermodynamics, but the
spatial landscape is invariant (cross-scale correlations >= 0.975) and
the MutL-AR association -- about 1 degree C lower median Tm and ~2.7
percentage points lower GC inside peaks -- is reproduced at every window
size (all p < 1e-12), so the conclusions do not hinge on the 200 bp
choice.

The workflow covers five steps:

1. **Build a BSgenome data package** from the NCBI assembly accession.
2. **Generate non-overlapping genomic windows** with `make_genomiccoord()`.
3. **Compute Tm and GC content** for all windows with `tm_calculate()`.
4. **Visualise** the genome-wide Tm profile with `plot_genome_track()` —
   circular plots, linear plots, zoom views, and multi-omics overlays.
5. **Statistical testing** — compare Tm inside MutL-AR peaks versus the rest
   of the genome with `compare_groups()`.

---

# Prerequisites

## R packages

```{r packages}
library(TmCalculator)
library(BSgenome)
library(GenomicRanges)
```

## Step 1 — Building the *E. coli* BSgenome data package {#build-bsgenome}

*TmCalculator* retrieves sequences from BSgenome data packages. A prebuilt
*E. coli* package is available on Bioconductor
(`BSgenome.Ecoli.NCBI.20080805`), but it is an older multi-strain snapshot
and does not provide the RefSeq assembly used here
(GCF_000005845.2, ASM584v2, sequence U00096.3). Because the MutL-AR peaks,
microsatellite, cruciform, ssDNA, and GATC tracks are all defined against
this assembly, we forge a matching BSgenome package locally from the NCBI
RefSeq sequence using **BSgenomeForge**, ensuring that genome sequence and
feature coordinates share a single reference. This step only needs to be run
once.

```{r forge, eval=FALSE}
library(BSgenomeForge)

forgeBSgenomeDataPkgFromNCBI(
  assembly_accession = "GCF_000005845.2",
  pkg_maintainer     = "Junhui Li <ljh.biostat@gmail.com>",
  destdir            = "."
)

install.packages(
  "./BSgenome.Ecoli.NCBI.ASM584v2",
  repos = NULL,
  type  = "source"
)
```

During *TmCalculator* package development and vignette building (e.g. `R CMD
check`), we cannot call `forgeBSgenomeDataPkgFromNCBI()` in the vignette
environment, so the chunks below install the pre-forged
`BSgenome.Ecoli.NCBI.ASM584v2` package from GitHub instead.

```{r setup-ecoli-bsgenome, message=FALSE, warning=FALSE}
ecoli_pkg   <- "BSgenome.Ecoli.NCBI.ASM584v2"
genome_obj  <- "Ecoli"   # BSgenomeObjname in DESCRIPTION; not the package name

.ecoli_genome_ready <- function() {
  if (!requireNamespace(ecoli_pkg, quietly = TRUE)) return(FALSE)
  exists(genome_obj, envir = asNamespace(ecoli_pkg), inherits = FALSE)
}

if (!requireNamespace(ecoli_pkg, quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) {
    utils::install.packages("remotes", repos = "https://cloud.r-project.org")
  }
  remotes::install_github(
    "JunhuiLi1017/BSgenome.Ecoli.NCBI.ASM584v2",
    upgrade = "never",
    quiet = TRUE
  )
}

if (!.ecoli_genome_ready()) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    utils::install.packages("BiocManager", repos = "https://cloud.r-project.org")
  }
  if (!requireNamespace("BSgenomeForge", quietly = TRUE)) {
    BiocManager::install("BSgenomeForge", ask = FALSE, update = FALSE)
  }
  pkgdir <- BSgenomeForge::forgeBSgenomeDataPkgFromNCBI(
    assembly_accession = "GCF_000005845.2",
    pkg_maintainer     = "Junhui Li <ljh.biostat@gmail.com>",
    destdir            = tempdir()
  )
  utils::install.packages(pkgdir, repos = NULL, type = "source", quiet = TRUE)
  if (ecoli_pkg %in% loadedNamespaces()) {
    unloadNamespace(ecoli_pkg)
  }
}

if (!.ecoli_genome_ready()) {
  stop(
    "Could not load genome object '", genome_obj, "' from package '", ecoli_pkg, "'.\n",
    "Run the manual 'forge' chunk above, or forgeBSgenomeDataPkgFromNCBI() locally.",
    call. = FALSE
  )
}
```

Load the genome (object `Ecoli`; package `BSgenome.Ecoli.NCBI.ASM584v2` for coordinates):

```{r load-genome}
ecoli_pkg  <- "BSgenome.Ecoli.NCBI.ASM584v2"
genome_obj <- "Ecoli"

suppressPackageStartupMessages(library(ecoli_pkg, character.only = TRUE))
genome      <- base::get(genome_obj, envir = asNamespace(ecoli_pkg))
genome_name <- ecoli_pkg
chr_name    <- "U00096.3"
chr_length  <- length(genome[[chr_name]])

cat("Chromosome:", chr_name, "\n")
cat("Length:    ", format(chr_length, big.mark = ","), "bp\n")
```

---

# Step 2 — Generate Genomic Windows

`make_genomiccoord()` tiles the chromosome into non-overlapping 200 bp bins.

```{r make-coord}
runtime0 <- system.time({
  bins_gc <- make_genomiccoord(
    bsgenome    = genome_name,
    chromosomes = chr_name,
    window      = 200L,
    slide       = 200L,
    start       = 1,
    end         = chr_length,
    strand      = "+"
  )
})

cat("Total windows:", length(bins_gc), "\n")
cat(sprintf("Window generation: %.2f s (elapsed)\n", runtime0["elapsed"]))
```

Resolve coordinates against the BSgenome package:

```{r coor-to-gr}
input_new <- list(pkg_name = genome_name, seq = bins_gc)
runtime1 <- system.time({
  gr_batch <- to_genomic_ranges_fast(input_new)
})

cat(sprintf(
  "Coordinate resolution: %.2f s (elapsed)\n",
  runtime1["elapsed"]
))
```

---

# Step 3 — Compute Genome-wide Tm

We use nearest-neighbour thermodynamics with the Breslauer *et al.* (1986)
parameter set at 50 mM Na+, matching the Methods of Hasenauer *et al.*
(2025), who computed Tm with this package and the same parameters.

```{r tm-calculate}
runtime2 <- system.time({
  tm_ASM584v2 <- tm_calculate(
    gr_batch,
    method   = "tm_nn",
    nn_table = "DNA_NN_Breslauer_1986",
    Na       = 50            # mM; standard PCR-like conditions
  )
})

cat(sprintf(
  "Tm calculation: %.2f s (elapsed) for %s windows\n",
  runtime2["elapsed"],
  format(length(bins_gc), big.mark = ",")
))

Tm <- as.data.frame(tm_ASM584v2$gr[, c("Tm", "GC")])
summary(Tm[, c("Tm", "GC")])
```

---

# Step 4 — Visualise with `plot_genome_track()`

`plot_genome_track()` is the unified plotting function in TmCalculator. It
supports both **linear** (karyoploteR-based) and **circular** (base R graphics)
layouts from the same track list, with features including ideogram tracks,
per-track highlights, proportional track heights, multi-region zoom, and
customisable legends.

## Assemble the track list

We overlay the Tm/GC profile with the Hasenauer *et al.* (2025) multi-omics
layers from `ecoli_rep_hotspots`:

| Track | Description |
|-------|-------------|
| MutL-AR | MutL ChIP-seq peaks marking replication-error / mismatch-repair hotspots |
| Microsatellites | Tandem-repeat (mononucleotide-repeat) density per 1 kb bin |
| Cruciform | Cruciform-forming sequence density per 1 kb bin |
| ssDNA | Single-stranded DNA regions enriched near error hotspots |
| GATC sites | Dam methylation-site (5'-GATC-3') density per 1 kb bin |

```{r track-list}
# Reference labels: replication origin (ori) and terminus (dif)
label <- data.frame(
  seqnames = genome_name,
  start    = c(3925804, 1590777),
  end      = c(3925804, 1590777),
  label    = c("ori", "dif")
)
data(ecoli_rep_hotspots)
tracks <- list(
  # Ideogram: MutL-AR peaks shown inside the chromosome bar
  list(type = "rect", data = ecoli_rep_hotspots$all_peaks_IP_mutH,
       col = "#2C3E50", bg.col = "grey", name = "MutL-AR",
       legend_font_col = "#2C3E50", ideogram = TRUE, height = 0.5),

  # Sequence thermodynamics
  list(type = "line", data = Tm, value_col = "GC",
       name = "GC content", col = "#4A90E2",
       legend_font_col = "#4A90E2"),
  list(type = "line", data = Tm, value_col = "Tm",
       name = "Melting temp", col = "#E06666",
       legend_font_col = "#E06666", height = 2),

  # Repeat / structural features
  list(type = "line", data = ecoli_rep_hotspots$bins_rep,
       value_col = "count", name = "Microsatellites", col = "#2ECC71",
       legend_font_col = "#2ECC71"),
  list(type = "line", data = ecoli_rep_hotspots$bins_cru,
       value_col = "count", name = "Cruciform", col = "#3B3E6B",
       legend_font_col = "#3B3E6B"),

  # ssDNA regions
  list(data = ecoli_rep_hotspots$ssdna, name = "ssDNA",
       col = "#8E44AD", legend_font_col = "#8E44AD"),

  # GATC methylation sites
  list(type = "line", data = ecoli_rep_hotspots$bins_gatc,
       value_col = "count", name = "GATC sites", col = "#D35400",
       legend_font_col = "#D35400"),

  # Global highlight: translucent bands at MutL-AR peaks across all tracks
  list(type = "highlight", data = ecoli_rep_hotspots$all_peaks_IP_mutH,
       col = "#F1C40F", alpha = 0.18)
)
```

## Circular genome map

```{r circular-full, fig.width=8, fig.height=8, fig.cap="Circular genome map of E. coli K-12 MG1655. Concentric rings from outside in: MutL-AR peaks (grey ideogram), GC content, melting temperature, microsatellite density, cruciform sequences, ssDNA regions, and GATC site density. Yellow highlight bands mark MutL-AR peak regions."}
plot_genome_track(
  genome_name = genome_name,
  genome_size = chr_length,
  track_list  = tracks,
  circular    = TRUE,
  label       = label
)
```

## Linear genome map

The same `tracks` list works for a linear karyoploteR layout — simply omit
`circular = TRUE`. The ideogram track is drawn inside the chromosome bar;
all other tracks are stacked as horizontal panels.

```{r linear-full, fig.width=10, fig.height=5, fig.cap="Linear genome view of E. coli K-12 MG1655. MutL-AR peaks are drawn inside the chromosome ideogram bar."}
plot_genome_track(
  genome_name = genome_name,
  genome_size = chr_length,
  track_list  = tracks
)
```

## Zoom — single region

The `zoom` parameter accepts a character string (e.g. `"chr:start-end"`) or a
GRanges object. In linear mode, the karyoploteR view is restricted to that
region; in circular mode, only data overlapping the region is drawn.

```{r zoom-single, fig.width=10, fig.height=5, fig.cap="Zoomed linear view of the 1.0-2.0 Mb region."}
plot_genome_track(
  genome_name = genome_name,
  genome_size = chr_length,
  track_list  = tracks,
  zoom        = c("U00096.3:100000-500000")
)
```

## Zoom — multiple regions

Pass a character vector to `zoom` to view several disjoint regions at once.
In linear mode each region is drawn as a separate stacked panel; in circular
mode the regions are concatenated around the circle with small gaps between
them.

```{r zoom-multi-linear, fig.width=10, fig.height=8, fig.cap="Two zoomed regions displayed as stacked linear panels."}
plot_genome_track(
  genome_name = genome_name,
  genome_size = chr_length,
  track_list  = tracks,
  circular    = TRUE,
  zoom        = c("U00096.3:100000-500000",
                  "U00096.3:3600000-4500000")
)
```

## Circular canvas panning

Use `canvas.xlim` and `canvas.ylim` to pan and magnify a portion of the
circular plot. The `circle.margin` parameter controls whitespace around the
plot.

```{r circular-pan, fig.width=7, fig.height=7, fig.cap="Panned circular view showing the upper-right quadrant of the E. coli chromosome."}
plot_genome_track(
  genome_name   = genome_name,
  genome_size   = chr_length,
  track_list    = tracks,
  circular      = TRUE,
  canvas.xlim   = c(0.5, 1),
  canvas.ylim   = c(0,   1),
  circle.margin = c(0.05, 0.05)
)
```

## Per-track highlights

Individual tracks can carry a `highlight` field to draw coloured bands within
that track only. This is useful for marking specific regions of interest on a
per-track basis:

```{r per-track-highlight, fig.width=8, fig.height=8, fig.cap="Circular plot with per-track highlight bands on the Microsatellites track."}
tracks_hl <- tracks
tracks_hl[[4]]$highlight <- list(
  data  = ecoli_rep_hotspots$bins_rep[1100:1200, ],
  col   = "black",
  alpha = 0.12
)

plot_genome_track(
  genome_name = genome_name,
  genome_size = chr_length,
  track_list  = tracks_hl,
  circular    = TRUE
)
```

---

# Step 5 — Statistical Testing

We test whether Tm values inside MutL-AR peaks differ significantly from the
rest of the genome.

### Build MutL-AR peak GRanges

```{r build-mutH-peaks}
mutH_peaks <- GRanges(
  seqnames = ecoli_rep_hotspots$all_peaks_IP_mutH$chr,
  ranges   = IRanges(start = ecoli_rep_hotspots$all_peaks_IP_mutH$start,
                     end   = ecoli_rep_hotspots$all_peaks_IP_mutH$end)
)
seqlevels(mutH_peaks) <- "U00096.3"
```

### Annotate Tm tiles with peak membership

```{r annotate-mutH-peaks}
mutH_peaks$peak_id <- paste0("mutH_", seq_along(mutH_peaks))

tm_annot <- integrate_granges(
  gr_tm          = tm_ASM584v2$gr,
  gr_features    = mutH_peaks,
  strategy       = "overlap",
  feature_cols   = "peak_id",
  keep_unmatched = TRUE
)

tm_annot$in_mutH <- ifelse(is.na(tm_annot$peak_id), "non_peak", "peak")
table(tm_annot$in_mutH)
```

### Wilcoxon test on Tm and GC

```{r wilcoxon}
res <- compare_groups(
  gr          = tm_annot,
  target      = c("Tm", "GC"),
  method      = "wilcoxon",
  group       = "in_mutH",
  alternative = "greater",
  posthoc     = FALSE
)
res$results
res$summary
```

---

# Interpreting the Results

MutL-AR-associated windows have lower Tm and lower GC content than the rest
of the genome (p < 0.001 for both; median Tm about 1 degree C lower and
mean GC 0.48 vs 0.51 inside peaks), consistent with the enrichment of
low-thermal-stability, AT-rich sequence reported by Hasenauer *et al.* MutL-AR peaks cluster near the
replication terminus, where replication forks converge and mismatch density is
highest. This spatial coincidence with locally elevated microsatellite density
and cruciform-forming sequences is consistent with the replication stress
model of MMR recruitment.

GATC methylation sites are distributed genome-wide but show local density
fluctuations that partially anti-correlate with GC content, reflecting the
sequence context requirements for Dam methyltransferase (5'-GATC-3').

---

# Window-Size Sensitivity Analysis

The 200 bp window used above is a design compromise, not a constraint of
the data: the external annotation layers have their own native
resolutions (1 kb bins for the microsatellite, cruciform and GATC density
tracks; variable-width intervals for MutL-AR peaks and ssDNA regions),
and both `plot_genome_track()` and the overlap-based statistics operate
across mixed resolutions. Here we recompute the Tm/GC profile at 50, 100,
200 and 500 bp and ask two questions: does the window size change the
thermodynamic landscape, and does it change the biological conclusion?

```{r window-sensitivity}
window_sizes <- c(50L, 100L, 200L, 500L)

sens <- lapply(window_sizes, function(w) {
  bins_w <- make_genomiccoord(
    bsgenome = genome_name, chromosomes = chr_name,
    window = w, slide = w, start = 1, end = chr_length,
    strand = "+", verbose = FALSE
  )
  gr_w <- to_genomic_ranges_fast(list(pkg_name = genome_name, seq = bins_w))
  tm_w <- tm_calculate(gr_w, method = "tm_nn",
                       nn_table = "DNA_NN_Breslauer_1986", Na = 50)$gr

  ann <- integrate_granges(gr_tm = tm_w, gr_features = mutH_peaks,
                           strategy = "overlap", feature_cols = "peak_id",
                           keep_unmatched = TRUE)
  ann$in_mutH <- ifelse(is.na(ann$peak_id), "non_peak", "peak")
  cg <- compare_groups(gr = ann, target = c("Tm", "GC"),
                       method = "wilcoxon", group = "in_mutH",
                       alternative = "greater", posthoc = FALSE)
  list(gr = tm_w, ann = ann, test = cg)
})
names(sens) <- paste0("w", window_sizes)
```

### Distributional stability

```{r sensitivity-distribution}
dist_tbl <- do.call(rbind, lapply(seq_along(window_sizes), function(i) {
  g <- sens[[i]]$gr
  data.frame(
    window_bp = window_sizes[i],
    n_windows = length(g),
    Tm_mean   = mean(g$Tm, na.rm = TRUE),
    Tm_sd     = stats::sd(g$Tm, na.rm = TRUE),
    Tm_IQR    = stats::IQR(g$Tm, na.rm = TRUE),
    GC_mean   = mean(g$GC, na.rm = TRUE)
  )
}))
knitr::kable(dist_tbl, digits = 3,
             caption = "Tm/GC distribution by window size.")
```

Two systematic effects are visible, both expected. First, mean Tm rises
with window length (66.4 to 78.9 degrees C from 50 to 500 bp): duplex
stability increases with length in the nearest-neighbor model, so
absolute Tm values are only comparable *within* one window size. Second,
shorter windows widen the distribution (SD 4.13 at 50 bp vs 2.54 at
500 bp) -- window size acts as a smoothing bandwidth. The mean GC is
invariant (0.508 throughout), confirming that the composition landscape
itself does not depend on the tiling.

### Cross-scale agreement

We aggregate each finer profile onto the 500 bp grid (mean Tm per bin)
and correlate it with the directly computed 500 bp profile:

```{r sensitivity-correlation}
gr500  <- sens[["w500"]]$gr
bin500 <- (start(gr500) - 1L) %/% 500L

cor_tbl <- vapply(c("w50", "w100", "w200"), function(k) {
  g   <- sens[[k]]$gr
  agg <- tapply(g$Tm, (start(g) - 1L) %/% 500L, mean, na.rm = TRUE)
  stats::cor(agg[match(bin500, as.integer(names(agg)))], gr500$Tm,
             use = "complete.obs")
}, numeric(1))
round(cor_tbl, 3)
```

Aggregated onto a common 500 bp grid, the profiles are nearly
interchangeable (r = 0.998, 0.999 and 0.975 for 50, 100 and 200 bp;
the 200 bp grid does not nest evenly into 500 bp bins, which blurs its
aggregate slightly). Window size rescales and smooths the profile but
preserves the spatial landscape.

### Robustness of the MutL-AR association

Finally, the vignette's biological conclusion -- that Tm and GC differ
between MutL-AR peak windows and the genomic background -- is re-tested
at every window size:

```{r sensitivity-tests}
## Per-window-size test statistics
test_results <- do.call(rbind, lapply(seq_along(window_sizes), function(i) {
  data.frame(window_bp = window_sizes[i], sens[[i]]$test$results)
}))
test_results

## Per-window-size group summaries, plus the median Tm effect size
test_summary <- do.call(rbind, lapply(seq_along(window_sizes), function(i) {
  ann <- sens[[i]]$ann
  med <- tapply(ann$Tm, ann$in_mutH, stats::median, na.rm = TRUE)
  data.frame(window_bp   = window_sizes[i],
             n_peak      = sum(ann$in_mutH == "peak"),
             Tm_med_diff = unname(med["peak"] - med["non_peak"]),
             sens[[i]]$test$summary)
}))
test_summary
```

### Multi-scale tracks alongside the multi-omics layers

Because every track keeps its native resolution, the four Tm/GC profiles
can be drawn as concentric layers of one integrated figure, together
with the annotation data:

```{r sensitivity-tracks, fig.width=8, fig.height=8, fig.cap="Multi-scale integration. Concentric rings from outside in: MutL-AR peaks (ideogram), GC content at 50/100/200/500 bp (blues), Tm at 50/100/200/500 bp (reds), and microsatellite density (green). Yellow bands mark MutL-AR peaks."}
## darkest = finest window (50 bp), lightest = coarsest (500 bp)
scale_cols_gc <- c("#1A5276", "#2E86C1", "#5DADE2", "#AED6F1")
scale_cols_tm <- c("#7B241C", "#CB4335", "#EC7063", "#F5B7B1")

sens_dfs <- lapply(sens, function(x) as.data.frame(x$gr[, c("Tm", "GC")]))

tracks_scale <- c(
  list(list(type = "rect", data = ecoli_rep_hotspots$all_peaks_IP_mutH,
            col = "#2C3E50", bg.col = "grey", name = "MutL-AR",
            legend_font_col = "#2C3E50", ideogram = TRUE, height = 0.5)),
  lapply(seq_along(window_sizes), function(i)
    list(type = "line", data = sens_dfs[[i]], value_col = "GC",
         name = paste0("GC ", window_sizes[i], " bp"),
         col = scale_cols_gc[i], legend_font_col = scale_cols_gc[i])),
  lapply(seq_along(window_sizes), function(i)
    list(type = "line", data = sens_dfs[[i]], value_col = "Tm",
         name = paste0("Tm ", window_sizes[i], " bp"),
         col = scale_cols_tm[i], legend_font_col = scale_cols_tm[i])),
  list(list(type = "line", data = ecoli_rep_hotspots$bins_rep,
            value_col = "count", name = "Microsatellites", col = "#2ECC71",
            legend_font_col = "#2ECC71"),
       list(type = "highlight",
            data = ecoli_rep_hotspots$all_peaks_IP_mutH,
            col = "#F1C40F", alpha = 0.18))
)

plot_genome_track(
  genome_name = genome_name,
  genome_size = chr_length,
  track_list  = tracks_scale,
  circular    = TRUE,
  label       = label
)
```

At whole-genome scale the four bandwidths look nearly identical -- the
landscape is invariant. The difference window size makes is local
smoothing, best seen in a zoomed view:

```{r sensitivity-zoom, fig.width=10, fig.height=7, fig.cap="Zoomed 1.5-1.7 Mb region. The 50 bp profile resolves single-window fluctuations that the 500 bp profile averages away; all four trace the same underlying landscape."}
plot_genome_track(
  genome_name = genome_name,
  genome_size = chr_length,
  track_list  = tracks_scale,
  zoom        = "U00096.3:1500000-1700000"
)
```

The association is reproduced at every scale: MutL-AR windows are about
1 degree C lower in median Tm than the background (-1.28, -1.02, -1.05
and -0.96 degrees C at 50, 100, 200 and 500 bp) and about 2.7 percentage
points lower in GC (0.482 vs 0.509), with all eight Wilcoxon tests
overwhelmingly significant (p from 1.5e-62 at 50 bp to 1.4e-12 at
500 bp; p shrinks with n, as expected, while the effect size is
essentially constant). The biological conclusion therefore does not
depend on the window choice. The primary analysis uses 200 bp to match
Hasenauer *et al.* (2025), whose Methods compute GC content and melting
temperature in 200 bp bins (using TmCalculator) and bin the ChIP-seq
coverage at the same 200 bp resolution.

---

# Computational Performance

On a standard desktop computer (single core, R 4.4, compiled Rcpp nearest-neighbor core):

| Step | Function | Time (elapsed) |
|------|----------|---------------|
| Window generation | `make_genomiccoord()` | ~`r sprintf("%.2f", runtime0["elapsed"])` s |
| Sequence extraction | `to_genomic_ranges_fast()` | ~`r sprintf("%.2f", runtime1["elapsed"])` s |
| Tm calculation (23,208 windows) | `tm_calculate()` | ~`r sprintf("%.2f", runtime2["elapsed"])` s |
| **Total** | | **~`r sprintf("%.2f", runtime0["elapsed"] + runtime1["elapsed"] + runtime2["elapsed"])` s** |

For human-scale analyses (non-overlapping 200 bp windows across the ~3.1 Gb
haploid genome; 14.7 M windows after excluding assembly gaps), run the
*entire pipeline* -- window building, sequence extraction and
`tm_calculate()` -- inside one `BiocParallel::SnowParam()` worker per
chromosome, keeping `tm_calculate()` itself at its serial default. On a
6-core/16 GB laptop this completes all of hg38 in just under 5 minutes
with 5 workers. Peak memory per worker is governed by its largest
chromosome (~4 GB for chr1), not by the total window count. Measured
timings, worker-count guidance and ready-to-run code are in the companion
vignette *Runtime and Memory for Genome-Scale Tm Profiling: an hg38
Benchmark Guide*.

---

# Session Information

```{r session-info}
sessionInfo()
```
