---
title: "Runtime and Memory for Genome-Scale Tm Profiling: an hg38 Benchmark Guide"
author: "Junhui Li, Lihua Julie Zhu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Runtime and Memory for Genome-Scale Tm Profiling: an hg38 Benchmark Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,   # requires BSgenome.Hsapiens.UCSC.hg38 and long runtimes
  message = FALSE,
  warning = FALSE
)
```

# Introduction

This vignette answers two practical questions that come up as soon as Tm
profiling is scaled from a bacterial genome to the human genome (hg38):

1. For a **single chromosome** (chr1), how long does the pipeline take, how
   much memory does it use, and why does supplying a multi-worker
   `BPPARAM` to `tm_calculate()` **not** make it faster?
2. For the **whole genome**, how should the work be parallelized, what are
   sensible parameter settings, and what runtime and memory should be
   budgeted?

All timings below were measured on a 6-core (12-thread) Intel MacBook Pro
with 16 GB RAM, R 4.4, macOS, with the package's compiled (Rcpp)
nearest-neighbor core. Absolute numbers will differ on other machines; the
*structure* of the results — which parts scale, which parts do not — will
not. None of the chunks are evaluated when the vignette is built; run them
interactively to reproduce the measurements on your own hardware.

```{r libraries}
library(TmCalculator)
library(BiocParallel)

pkg <- "BSgenome.Hsapiens.UCSC.hg38"
suppressPackageStartupMessages(library(pkg, character.only = TRUE))
genome <- get(pkg, envir = asNamespace(pkg))
```

# Measuring time and memory

Three tools cover almost every need. Wall-clock time comes from
`system.time()` (read the `elapsed` column; for parallel runs `user` and
`system` only reflect the manager process). The peak of R-managed memory
over a code block comes from the `gc(reset = TRUE)` idiom. The true
resident size of the current process — including C++ allocations that R's
GC does not see — comes from `ps::ps_memory_info()`.

```{r measuring}
base::gc(reset = TRUE)                # reset the "max used" high-water mark
## ... run the code to be measured ...
base::gc()                            # read peak from the "max used" column

ps::ps_memory_info()[["rss"]] / 1e9   # current process resident set, GB
```

Note the `base::` prefix: TmCalculator exports a `gc()` function of its
own (GC *content* of a sequence), which masks R's garbage collector once
the package is attached.

For `SnowParam()` workers the manager cannot see worker memory at all;
record it inside the worker function and return it with the result (this
pattern is used in the whole-genome section below).

# A ten-second warm-up: chr21

Before committing to chr1, verify your setup on chr21 — at 46.7 Mb it is
the smallest hg38 autosome and the whole pipeline finishes in about ten
seconds. It is also a deliberately instructive choice: chr21 *begins*
with a ~5 Mb assembly gap of `N` bases, so it exercises the N-trimming
machinery that most chromosomes only touch lightly.

```{r chr21-warmup}
chr_len21 <- GenomeInfoDb::seqlengths(genome)[["chr21"]]

t21 <- system.time({
  bins21 <- make_genomiccoord(bsgenome = pkg, chromosomes = "chr21",
                              window = 200L, slide = 200L,
                              start = 1, end = chr_len21, strand = "+")
  gr21   <- to_genomic_ranges_fast(list(pkg_name = pkg, seq = bins21),
                                   method = "preload_chr")
  tm21   <- tm_calculate(gr21, method = "tm_nn",
                         nn_table = "DNA_NN_SantaLucia_2004", Na = 50)
})
t21["elapsed"]        # ~10-15 s cold, ~6 s warm, on the test machine
head(tm21$gr)
```

Watch the `make_genomiccoord()` messages: N-trimming reports an effective
range starting at position 5,010,001 — the leading 5 Mb of `N`s is
detected in a single in-memory scan and excluded before any windows are
built, leaving 208,449 windows instead of 233,549. The first run in a
session pays one-time costs (reading the genome from disk, loading the
thermodynamic tables); subsequent runs are roughly twice as fast. If this
warm-up runs cleanly, your BSgenome installation, compiled core and
memory headroom are all in working order, and the chr1 numbers below
should reproduce within noise.

One benchmarking caveat that bites easily: `devtools::load_all()`
compiles the C++ core at `-O0` (a debug build, ~6x slower) and leaves a
mixed session state even after reinstalling. Always benchmark in a fresh
session with a plain `library(TmCalculator)`; confirm with
`"devtools_shims" %in% search()` returning `FALSE`.

# Part 1: a single chromosome (chr1)

## The pipeline and its serial cost

chr1 is 248,956,422 bp. Non-overlapping 200 bp windows after N-trimming
give 1,244,682 windows.

```{r chr1-serial}
chr_len <- GenomeInfoDb::seqlengths(genome)[["chr1"]]

t_coord <- system.time({
  bins <- make_genomiccoord(bsgenome = pkg, chromosomes = "chr1",
                            window = 200L, slide = 200L,
                            start = 1, end = chr_len, strand = "+")
})

t_extract <- system.time({
  gr_batch <- to_genomic_ranges_fast(list(pkg_name = pkg, seq = bins),
                                     method = "preload_chr")
})

base::gc(reset = TRUE)
t_tm <- system.time({
  tm_chr1 <- tm_calculate(gr_batch, method = "tm_nn",
                          nn_table = "DNA_NN_SantaLucia_2004", Na = 50)
})
base::gc()   # "max used" = peak R memory during the Tm step

rbind(t_coord, t_extract, t_tm)[, "elapsed"]
## t_coord t_extract      t_tm
##     5.5      15.4      30.8     (freshly booted, otherwise idle machine)
```

Measured on the test machine (`R --vanilla`, freshly booted, no other
applications):

| step | function | elapsed | notes |
|---|---|---|---|
| window coordinates | `make_genomiccoord()` | ~5.5 s | N-end scan loads the 249 Mb chromosome once |
| sequence extraction | `to_genomic_ranges_fast(method = "preload_chr")` | ~15 s | `extractAt()` itself is ~4 s; the rest is complement generation, the GC column and `GRanges` assembly |
| Tm calculation | `tm_calculate(method = "tm_nn")` | ~31 s | compiled core, 1.24M windows |
| **total** | | **~52 s** | |

On a machine with normal background load the same steps run 20-35%
slower (we measured 7.3 / 19.3 / 41.5 s in an ordinary desktop session);
comparisons within one section below were always measured in a single
session.

Memory: `base::gc()` reported a peak of ~1.9 GB of R-managed heap during
the Tm step, and a dedicated worker process running the whole chr1
pipeline reaches **~3.5-4 GB resident** (sequences, complements, a
character copy handed to the compiled core, `GRanges` metadata, plus
loaded packages). A useful planning figure: **budget ~4 GB per
chr1-sized pipeline**, tapering to well under 2 GB for small
chromosomes.

Two extraction details matter at this scale. Pass the input as
`list(pkg_name = ..., seq = ...)` so the genome package is loaded once,
and use `method = "preload_chr"` for dense tiling of few chromosomes
(`getSeq()`-based `"vectorized"` is better for windows scattered across
many chromosomes).

## Why `BPPARAM` does not accelerate a single chromosome

`tm_calculate()` accepts a `BPPARAM` and it does exactly what it
advertises — yet on chr1 it does not help:

```{r chr1-parallel}
system.time({
  tm_serial <- tm_calculate(gr_batch, method = "tm_nn", Na = 50)
})
## elapsed ~ 30 s

system.time({
  tm_snow <- tm_calculate(gr_batch, method = "tm_nn", Na = 50,
                          BPPARAM = SnowParam(workers = 5))
})
## elapsed ~ 55 s on the idle machine (33-80 s across repeated sessions)
## -- NEVER faster than serial
```

The serial time is stable across sessions; the parallel time is not,
because its added costs (spawning workers, serializing half a gigabyte
of sequence, loading packages in each worker) are highly sensitive to
machine state. The decomposition explains why even its best case cannot
win. Of the ~30 s serial runtime, only the per-window Tm loop (~9 s with
the compiled core) is divided among workers. The rest is inherently
serial or is *new* cost created by parallelization:

| component | ~time | scales with workers? |
|---|---|---|
| serial preprocessing (N filtering, coercion, result assembly) | ~21 s | no — runs once in the manager |
| shipping ~0.5 GB of sequence to workers and results back | ~5 s | no — serialization is serial |
| worker startup + package loading | ~3+ s | fixed cost, *added* by parallelism |
| compiled Tm loop | ~9 s / n workers | **yes** |

This is Amdahl's law with an unusually small parallel fraction (~30%):
even with infinitely many workers the floor is ~29 s — essentially the
serial runtime — and on a loaded machine the added costs balloon well
past it. With the pure-R core of previous releases the loop
dominated (minutes for an input of this size) and the same
`SnowParam(5)` call gave a 2.8x speedup; compiling the loop moved the
bottleneck into the serial part and made within-call parallelism
unprofitable. **With the compiled core, the serial default
(`SerialParam()`) is the fastest way to run inputs up to roughly a
million windows.**

A stronger warning applies to `MulticoreParam()` on large inputs. Forked
workers share the manager's memory copy-on-write, but R's garbage
collector *writes* to every object header it marks, so each worker's GC
forces the kernel to physically duplicate inherited pages. On a
chr1-scale input (measured with the pure-R core, where the effect is
easiest to see), `MulticoreParam(5)` ran *slower than serial* while
burning ~18 CPU-minutes of `system` time on page duplication. If you
parallelize at all, prefer `SnowParam()`.

# Part 2: the whole genome

## Parallelize at the chromosome level

The serial preprocessing that defeats within-call parallelism is itself
embarrassingly parallel *across chromosomes*: window building, extraction,
filtering and assembly for chr2 are independent of chr1. So for
genome-scale work, give each worker one whole chromosome and run the
entire pipeline — coordinates, extraction, `tm_calculate()` with the
serial default — inside the worker:

```{r genome-parallel}
chrs <- paste0("chr", c(1:22, "X", "Y"))
## Better: sort largest-first for load balance -- see "Choosing the
## worker count" below for the one-liner.
n_workers <- 5   # see "Choosing the worker count" below

runtime <- system.time({
  res_list <- bplapply(chrs, function(chr, pkg) {
    ## SnowParam workers are fresh R processes: load packages HERE,
    ## not in the manager session.
    suppressPackageStartupMessages(library(TmCalculator))
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))

    genome  <- get(pkg, envir = asNamespace(pkg))
    chr_len <- GenomeInfoDb::seqlengths(genome)[[chr]]

    bins <- make_genomiccoord(bsgenome = pkg, chromosomes = chr,
                              window = 200L, slide = 200L,
                              start = 1, end = chr_len, strand = "+",
                              verbose = FALSE)
    gr  <- to_genomic_ranges_fast(list(pkg_name = pkg, seq = bins),
                                  method = "preload_chr")
    out <- tm_calculate(gr, method = "tm_nn",
                        nn_table = "DNA_NN_SantaLucia_2004",
                        Na = 50)$gr        # serial inside the worker

    ## Drop sequence columns before returning: Tm/GC are what we keep,
    ## and this cuts per-chromosome serialization from ~500 MB to a few MB.
    out$sequence   <- NULL
    out$complement <- NULL

    ## Record this worker's peak resident memory (GB) for budgeting.
    attr(out, "worker_rss_gb") <- ps::ps_memory_info()[["rss"]] / 1e9
    out
  }, pkg = pkg, BPPARAM = SnowParam(workers = n_workers))

  tm_genome <- unlist(GenomicRanges::GRangesList(res_list))
})

runtime
sapply(res_list, attr, "worker_rss_gb")   # per-worker memory check
length(tm_genome)                         # 14,687,330 windows
summary(tm_genome$Tm)
## Min. 47.2  1st Qu. 68.9  Median 71.5  Mean 72.0  3rd Qu. 74.8  Max. 101.1
```

Measured on the test machine (all 24 chromosomes, 14,687,330 windows,
freshly booted, `R --vanilla`), sweeping the worker count:

| workers | elapsed | note |
|---|---|---|
| 3 | 387 s | |
| 4 | 356 s | |
| 5 | **295 s** | optimum = physical cores − 1 |
| 6 | 338 s | slower: core contention begins |
| 7 | 346 s | slower still |

So the whole genome takes **just under 5 minutes** at the optimum,
versus roughly 10 minutes run serially (estimated from the
per-chromosome costs) — about **2x**. Unlike within-call `BPPARAM`, this
scheme parallelizes *everything* — window building, extraction,
filtering and assembly, not just the Tm loop — and per-task
communication is a chromosome name in and a slim `GRanges` out, so
Amdahl's law has little serial residue to bite on. The speedup is
bounded near 2x on this laptop by load imbalance (chr1 alone takes
~50 s), disk contention while several workers read the genome file, and
the 16 GB memory ceiling. Per-worker resident memory in the sweep ranged
from ~0.3 GB (small chromosomes) to ~4.1 GB (chr1/chr2), consistent with
the ~4 GB planning figure from Part 1.

## Finer grain: segment-level parallelism

The same pattern works below the chromosome level, and it is the answer
to "so how *do* I parallelize a single chromosome?": ship coordinates,
not sequences. Each worker receives a `(start, end)` pair — a few dozen
bytes — and runs the entire pipeline on its own segment. Every row of
the Part 1 cost table becomes parallel except worker startup, because
the preprocessing happens inside the workers and the half-gigabyte
serialization never occurs.

```{r segment-parallel}
chr_len <- GenomeInfoDb::seqlengths(genome)[["chr1"]]

## Segment length must be a multiple of `slide` so the window grid stays
## aligned across segment boundaries (50 Mb = 250,000 x 200).
seg_starts <- seq(1, chr_len, by = 50e6)
seg <- data.frame(start = seg_starts,
                  end   = pmin(seg_starts + 50e6 - 1, chr_len))

t_seg <- system.time({
  res_seg <- bplapply(seq_len(nrow(seg)), function(i, seg, pkg) {
    suppressPackageStartupMessages(library(TmCalculator))
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))
    bins <- make_genomiccoord(bsgenome = pkg, chromosomes = "chr1",
                              window = 200L, slide = 200L,
                              start = seg$start[i], end = seg$end[i],
                              strand = "+", trim_N = "none",
                              verbose = FALSE)
    gr <- to_genomic_ranges_fast(list(pkg_name = pkg, seq = bins),
                                 method = "preload_chr")
    out <- tm_calculate(gr, method = "tm_nn",
                        nn_table = "DNA_NN_SantaLucia_2004", Na = 50)$gr
    out$sequence <- NULL; out$complement <- NULL
    out
  }, seg = seg, pkg = pkg, BPPARAM = SnowParam(workers = 5))

  tm_chr1_seg <- sort(unlist(GenomicRanges::GRangesList(res_seg)))
})

t_seg["elapsed"]      # 38 s measured, vs ~52 s for the serial FULL pipeline
length(tm_chr1_seg)   # 1,152,300
```

Mind the comparison scope: 38 s covers the *entire* pipeline (windows,
extraction, Tm), so the serial reference is the ~52 s pipeline total from
Part 1 — a **1.35x** speedup — not the 30 s Tm-only figure. Modest, but
it is the only single-chromosome parallelization that wins at all
(within-call `SnowParam` *lost* to serial above), and the result is
verifiably identical: 1,152,300 windows equals the serial run's
1,244,682 generated minus its 92,382 N-gap windows, exactly. That
equivalence is by construction — `trim_N = "none"` keeps the window grid
identical across segments (the "Skipped ... regions" warning from
`tm_calculate()` is expected; hg38 chr1 carries an ~18 Mb interior
centromeric gap), and the 50 Mb segment length is a multiple of `slide`.

The gain is capped well below worker-count because each worker pays
fixed costs the serial run pays once: loading packages and reading the
same 249 Mb chromosome to extract its segment (five workers contending
for the same file). Per-worker memory stays moderate (~1-2 GB, since
only a 50 Mb segment is windowed). For whole-genome runs this scheme is
the natural refinement anyway: segmenting the large chromosomes removes
the ~50 s chr1 straggler that limits the per-chromosome scheme's load
balance.

## Choosing the worker count

Each worker's peak memory is governed by its largest chromosome —
measured at **~4 GB for a chr1-sized task**, tapering to well under 1 GB
for the small chromosomes. Two constraints, take the minimum:

```{r workers}
mem_gb    <- 16                                     # your machine
cores     <- parallel::detectCores(logical = FALSE) # physical cores
n_workers <- min(cores - 1L, floor((mem_gb - 4) / 3))
n_workers
## 16 GB / 6 cores  -> 4 workers (safe); the sweep above found 5 fastest
## 32 GB / 8 cores  -> 7 workers
## 64 GB / 10 cores -> 9 workers
```

The worker sweep above pins down both halves of the formula. The core
bound is exact: the optimum was `cores - 1 = 5`, and adding a 6th or 7th
worker made the run *slower*, not faster. The memory term is
deliberately conservative: it assumes every worker holds a large
chromosome at once, which rarely happens, so on this 16 GB machine 5
workers (one more than the formula's 4) was in practice both safe and
fastest — but 4 is the defensible default for an unattended run. Do not
count hyperthreads: the compiled core is CPU-bound and logical cores add
little beyond scheduler overhead. And treat the memory bound as a hard
one — not every worker holds a large chromosome simultaneously (which is
why 5 workers survived on 16 GB), but when they do collide the system
swaps, and a swapping parallel run is *slower than serial* (we measured
19 GB of swap and a wedged machine in exactly this state, aggravated by
leftover workers from interrupted runs). Sorting chromosomes
largest-first also helps load balance, since the default scheduler deals
them out in order:

```{r sort-chrs}
sl   <- GenomeInfoDb::seqlengths(genome)
chrs <- names(sort(sl[paste0("chr", c(1:22, "X", "Y"))], decreasing = TRUE))
```

## Housekeeping

Four operational notes. First, benchmark in a session that did *not*
restore a saved workspace (`[Workspace loaded from ~/.RData]` at startup
means gigabytes of old objects are already resident); disable
"Restore .RData at startup" in RStudio or start `R --vanilla`. Second,
`SnowParam()` workers load the *installed* package: after
`devtools::load_all()` development changes, run `devtools::install()`
before benchmarking, and note that `load_all()` compiles C++ at `-O0` —
time only the installed build. The same applies to
`devtools::build_vignettes()`, which evaluates vignette code against a
`-O0` debug build (timings come out ~6x slow); set
`options(pkg.build_extra_flags = FALSE)` to disable the debug flags, or
rely on `R CMD build` / `devtools::check()`, which install with normal
optimization before running vignettes. Changing the flag option does
*not* trigger recompilation — stale `-O0` object files in `src/` are
reused as long as sources are unchanged — so run `pkgbuild::clean_dll()`
after changing it.
Third, if a parallel run errors or is interrupted, workers can linger
and hold gigabytes; check for orphaned `R` processes (Activity Monitor on
macOS, `ps aux | grep R` elsewhere) before rerunning. Fourth, worker
`message()` output is buffered until each task finishes — a silent
console does not mean a hung run; watch the worker processes' CPU
instead.

# Summary

For hg38 with the compiled nearest-neighbor core: run single chromosomes
serially (`BPPARAM` left at its default — within-call parallelism cannot
beat Amdahl's law when the parallel fraction is ~30%), and run the whole
genome as one chromosome per `SnowParam()` worker with the worker count
set by `min(physical cores - 1, memory budget)`. Measured on a
6-core/16 GB laptop: ~52 s and up to ~4 GB per chr1-sized
chromosome-worker, and 295 s (just under 5 minutes) for the full genome
(14.7 million windows) at the measured optimum of 5 workers, versus
~10 minutes serially.

# Session info

```{r sessioninfo, eval=TRUE}
sessionInfo()
```
