---
title: "Genome-Scale Tm Profiling on hg38: Inputs, Parallel Execution and Performance"
author: "Junhui Li, Lihua Julie Zhu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genome-Scale Tm Profiling on hg38: Inputs, Parallel Execution and Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,   # the hg38 chunks need BSgenome.Hsapiens.UCSC.hg38 and minutes
  message = FALSE,
  warning = FALSE,
  fig.retina = 1,
  dpi        = 72
)
```

# Introduction

`tm_calculate()` turns a genome, or any set of regions in it, into a melting
temperature profile: it tiles the regions into windows, fetches each
window's sequence, computes a Tm, and returns the lot as one object. The
same call takes sequences you already have, so there is one function to
learn; on a mammalian genome it spreads the work across processes.

This vignette answers three questions, in order:

1. **What can I give it?** A BSgenome package, a FASTA file, or sequences
   already in R, and any set of regions within them.
2. **How does it divide the work?** Into tasks, one region or one segment
   each, dispatched to workers.
3. **How many workers should I use, and how long will it take?** Measured on
   the human genome in two hardware environments.

**Before you start.** You need a BSgenome package for the genome you want to
profile; this vignette uses `BSgenome.Hsapiens.UCSC.hg38`. **BiocParallel**
is needed only if you want more than one worker. Neither is required for
`tm_calculate()` to run in a single process.

```{r install, eval=FALSE}
BiocManager::install(c("BSgenome.Hsapiens.UCSC.hg38", "BiocParallel"))
```

The hg38 chunks below are not evaluated when the vignette is built, since
they take minutes; the timing table and the figure are evaluated, and are
built from measurements that ship with the package.

# Quick start

The smallest genome-scale call needs a genome and a window width:

```{r quick-start}
library(TmCalculator)
hg38 <- "BSgenome.Hsapiens.UCSC.hg38"

tm <- tm_calculate(hg38, window = 200, slide = 200)$gr
tm
## GRanges object with 14687412 ranges and 2 metadata columns:
##       seqnames      ranges strand |        Tm        GC
```

`tm_calculate()` returns a `TmCalculator` object: `$gr` is the profile and
`$options` records the model and tiling it was produced with.

That tiles every standard chromosome into non-overlapping 200 bp windows
and computes a nearest-neighbour Tm for each, in a single process. The
window width has no default at genome scale, and is not meant to: it is the
resolution of the profile, so it belongs in the call rather than in a
default that quietly decides how many rows come back. Adding workers is one
more argument:

```{r quick-start-parallel}
library(BiocParallel)

tm <- tm_calculate(hg38, window = 200, slide = 200,
                   BPPARAM = SnowParam(workers = 5))$gr
```

Same windows, same Tm values, about a third of the time. How the genome is
divided changes the runtime, not the result.

Everything else is a refinement of those two calls: which regions, which
thermodynamic model, and how many workers.

# What can I give it?

## The sequence source

The first argument says where sequence comes from. It is given by *name*,
not as a loaded object, because each worker opens the source for itself.

```{r sources}
tm_calculate(hg38, window = 200)         # an installed BSgenome package
tm_calculate("contigs.fa.gz", window = 200)   # a FASTA file, gzip is fine
tm_calculate(oligos)                     # a character vector of sequences
```

A vector of sequences is staged as a temporary FASTA under `tmpdir` and
deleted on exit, so that the workers read it rather than receive it. That is
worth doing for a large set, because it moves window construction and result
assembly into the workers as well; for a handful of sequences the staging
and the worker start-up cost more than the calculation, and `tm_calculate()`
is the right call. On a cluster, set `tmpdir` to node-local scratch: the
default `tempdir()` is often a small partition.

## The regions

`regions` says what to take from the source, in whichever form is at hand.

```{r regions}
## Chromosomes or records, by name or by number. On a BSgenome the chr
## prefix is added or removed to match the genome, so the same code works on
## UCSC and on Ensembl; FASTA record names are matched exactly.
tm_calculate(hg38, regions = 1:22, window = 200)               # autosomes
tm_calculate(hg38, regions = paste0("chr", c(1:22, "X", "Y")), # no chrM
             window = 200)
tm_calculate("contigs.fa.gz", regions = c("contig_7", "contig_9"),
             window = 200)

## Coordinate intervals. Commas and scientific notation are accepted, so a
## number pasted out of a genome browser works as it stands.
tm_calculate(hg38, regions = c("chr1:1-10e6", "chrX:5,000,000-6,000,000"),
             window = 200)

## A mixture, when some chromosomes are wanted whole and others in part.
tm_calculate(hg38, regions = c(1:20, "chr21:1-10e6", "X", "Y"), window = 200)

## A GRanges, when the regions come from an annotation.
prom <- promoters(genes(TxDb.Hsapiens.UCSC.hg38.knownGene),
                  upstream = 1000, downstream = 500)
tm_calculate(hg38, regions = prom, window = 50, slide = 25)

## A GRanges carrying its own sequences is a source rather than a query, and
## then regions selects by overlap: a seqname takes every range on it, an
## interval takes the ranges it meets. Whole ranges come back, not clipped
## pieces of them, since their sequences are already fixed.
tm_calculate(probes_gr, regions = "chr7")
tm_calculate(probes_gr, regions = "chr7:1-1e6")

## window = NULL, the default, gives one window per region, which is what
## short records call for: a FASTA of array probes, primers or synthetic
## oligos returns one Tm per record. It is refused for a region over 1 Mb,
## where a single Tm would mean nothing.
tm_calculate("probes.fa", window = NULL)
tm_calculate(oligos, BPPARAM = SnowParam(5))
```

Three things are worth knowing before you trust the output.

*The default includes chrM.* With no `regions`, a BSgenome source covers
`GenomeInfoDb::standardChromosomes()`, which for GRCh38 is the 24 assembled
chromosomes **and** the mitochondrion. Name the chromosomes explicitly if
that does not belong in your profile. The sweep later in this vignette uses
`paste0("chr", c(1:22, "X", "Y"))` for exactly that reason, and its
14,687,330 windows are the 24 without chrM.

*Whole chromosomes are trimmed, named regions are not.* A region that names
a whole chromosome has its leading and trailing assembly gaps trimmed, since
a telomeric run of N carries no windows. A region given by coordinate is
tiled from the start you asked for. Windows containing N are dropped either
way.

*Overlapping regions are not merged.* They produce windows that appear twice
in the result, so `tm_calculate()` warns rather than double-counting them
quietly.

## Array probes

Array manifests are not read directly, since their layout is vendor and
version specific. Take the probe sequences out with the package that already
understands the manifest, `illuminaio`, `minfi` or `sesame` for Infinium
arrays, then pass them as a FASTA file or straight to `tm_calculate()`.

# How does it divide the work?

A **task** is the unit one worker owns from start to finish: it opens the
source, builds its own windows, fetches its own sequence and computes its
own Tm. Only a name and a coordinate pair cross between processes, so no
sequence is ever serialized. `unit` decides how tasks are cut.

```{r unit}
tm_calculate(hg38, unit = "segment", segment_size = 50e6,   # default: 73 tasks
           BPPARAM = SnowParam(workers = 5))
tm_calculate(hg38, unit = "region",                         # 24 tasks
           BPPARAM = SnowParam(workers = 5))
```

**Segments are the better default**, for two reasons. Chromosome 1 is an
indivisible task of 249 Mb, so with one task per chromosome the run cannot
finish before chromosome 1 does, however many workers are available. And a
worker holding a 50 Mb segment needs less memory than one holding a whole
chromosome. Segment boundaries fall on multiples of `slide`, so the window
grid, and therefore the result, is identical either way.

**Prefer `SnowParam()` to `MulticoreParam()`.** Forked workers share the
manager's memory copy-on-write, but R's garbage collector writes to every
object header it marks, so each worker's collection forces the kernel to
duplicate the inherited pages. On a genome-scale input that has been
measured running slower than a single process.

# How many workers, and how long?

Memory, not cores, is usually what limits the worker count, because each
worker holds the sequences of its own task.

The sweep below ships with the package: hg38 at 200 bp non-overlapping
windows, 14,687,330 windows, one to six workers, three repetitions per
configuration, on a six-core 16 GB laptop and on a cluster compute node with
six slots and a 48 GB memory limit.

```{r sweep-data, eval=TRUE}
# Read the shipped sweeps rather than transcribing numbers, so the table
# cannot drift from the measurements it describes. The two files do not
# carry the same columns, the cluster sweep also recording the host and CPU
# model it landed on, so both are cut to the four columns used here before
# they are stacked.
cols <- c("strategy", "n_workers", "wall_s", "max_rss_gb")
read_sweep <- function(file, env) {
  d <- utils::read.csv(system.file("extdata", file, package = "TmCalculator"),
                       stringsAsFactors = FALSE)
  d <- d[d$strategy == "segment", cols]      # the default unit of tm_calculate()
  d$env <- env
  d
}
raw <- rbind(read_sweep("bench_parallel_strategy.csv", "Laptop"),
             read_sweep("bench_parallel_cluster.csv",  "Compute node"))

sweep <- do.call(rbind, lapply(
  split(raw, list(raw$env, raw$n_workers), drop = TRUE),
  function(d) data.frame(
    Environment = d$env[1], Workers = d$n_workers[1],
    wall_s = median(d$wall_s), lo = min(d$wall_s), hi = max(d$wall_s),
    rss_gb = median(d$max_rss_gb), stringsAsFactors = FALSE)))

serial <- stats::setNames(sweep$wall_s[sweep$Workers == 1L],
                          sweep$Environment[sweep$Workers == 1L])
sweep$Speedup <- serial[sweep$Environment] / sweep$wall_s
sweep <- sweep[order(sweep$Environment != "Laptop", sweep$Workers), ]
```

```{r sweep-table, eval=TRUE}
knitr::kable(
  data.frame(
    Environment = sweep$Environment,
    Workers     = sweep$Workers,
    `Wall time (s)` = sprintf("%.1f [%.1f-%.1f]", sweep$wall_s, sweep$lo, sweep$hi),
    Speedup     = sprintf("%.2f", sweep$Speedup),
    `Peak RSS per worker (GB)` = sprintf("%.2f", sweep$rss_gb),
    check.names = FALSE),
  row.names = FALSE,
  caption = paste("Median of three repetitions, observed range in brackets.",
                  "Wall time includes worker start-up."))
```

```{r sweep-figure, eval=TRUE, fig.width=8, fig.height=3.6, fig.cap="Wall-clock time and peak memory per worker against worker count, on a six-core 16 GB laptop (solid, filled) and a compute node with a 48 GB limit (dashed, open). Points are medians of three repetitions; bars give the observed range."}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.4, 2.2, 0.8), mgp = c(2.6, 0.7, 0))
for (what in c("wall", "rss")) {
  ys <- if (what == "wall") sweep$wall_s else sweep$rss_gb
  plot(range(sweep$Workers), range(0, ys * 1.05), type = "n",
       xlab = "Workers", xaxt = "n", adj = 0,
       ylab = if (what == "wall") "Wall clock (s)" else "Peak resident size per worker (GB)",
       main = if (what == "wall") "A" else "B")
  axis(1, at = sort(unique(sweep$Workers)))
  for (e in unique(sweep$Environment)) {
    d <- sweep[sweep$Environment == e, ]
    y <- if (what == "wall") d$wall_s else d$rss_gb
    solid <- e == "Laptop"
    if (what == "wall")
      arrows(d$Workers, d$lo, d$Workers, d$hi, angle = 90, code = 3,
             length = 0.03, col = "grey40")
    lines(d$Workers, y, lty = if (solid) 1 else 2, col = "grey20")
    points(d$Workers, y, pch = if (solid) 19 else 1, col = "grey20")
  }
  if (what == "wall")
    legend("topright", c("Laptop, 16 GB", "Compute node, 48 GB"),
           lty = c(1, 2), pch = c(19, 1), bty = "n", cex = 0.85, col = "grey20")
}
par(op)
```

Two things in that table decide a worker count.

**More workers stop helping, and where depends on memory.** The laptop is
fastest at five workers, 189 s against 578.5 s in one process, and a sixth
worker makes the run *slower*. The node, with three times the memory,
improves all the way to six workers and 127 s. The turnover is memory
pressure rather than anything in the software, which is why the optimum
belongs to the machine and not to the package.

**Memory per worker falls as you add workers**, from 4.87 GB in the
one-worker run to 1.86 GB at five, because no worker holds more than one
50 Mb task at a time. That is what makes more workers affordable at all.

A rule that works on both machines:

> workers = min(physical cores - 1, (available GB - 4) / 2)

Confirm it on your own machine with a short sweep over one chromosome rather
than the whole genome. Note that worker start-up, roughly ten seconds per
call, is included in every timing, so a run that takes less than a minute
will understate the speedup:

```{r sweep-your-machine}
for (n in 1:6) {
  t <- system.time(
    tm_calculate(hg38, regions = "chr21", window = 200, slide = 200,
               method = "tm_nn", segment_size = 10e6,
               BPPARAM = SnowParam(workers = n), verbose = FALSE))
  cat(sprintf("%d workers: %.1f s\n", n, t[["elapsed"]]))
}
```

The sweeps above were produced by `bench_parallel_strategy.R` on the laptop
and by `bench_parallel_cluster.R`, with its LSF submission file
`bench_parallel_cluster.lsf`, on the compute node. All three are in
`system.file("scripts", package = "TmCalculator")`, and their output is the
two CSV files read here, `bench_parallel_strategy.csv` and
`bench_parallel_cluster.csv` in `extdata`, so re-running one on your own
hardware means replacing a file.

`bench_tm_calculate.R` in the same directory times `tm_calculate()` itself over
a range of worker counts, with `bench_tm_calculate.lsf` to submit it, and is
the one to reach for when sizing a new machine.

# What comes back

A `GRanges` with `Tm` and `GC` metadata columns in genomic order, ready for
`integrate_granges()`, `compare_groups()` and the plotting functions, and
for anything else that takes genomic intervals.

```{r downstream}
tm_annot <- integrate_granges(gr_tm = tm, gr_features = atac_peaks,
                              strategy = "overlap", weight = "overlap")
compare_groups(tm_annot, value_cols = "Tm", group_col = "class")
```

The sequence and complement columns are dropped by default, since they run
to roughly 500 MB per large chromosome; pass `keep_sequence = TRUE` if you
need them.

# One function, two ways of using it

`tm_calculate()` starts either from a source it can open, as above, or from
sequences you already hold:

```{r tm-calculate}
tm_calculate(c("ACGTGCTAGCTAGCTAGC", "GGCCATATATGCGC"), method = "tm_nn", Na = 50)
```

Given sequences and nothing else it does exactly what it has always done,
one Tm per sequence, by the shortest path through the function. Add
`regions`, `window` or `BPPARAM` and the profiling machinery engages.

What it will not do is divide the sequences of one region among workers.
With the compiled core the per-window loop is a minority of a call's cost;
window construction, sequence retrieval and result assembly run once, and
sending the sequences to workers costs more than the loop it divides.
Measured on chromosome 1, splitting one call across five workers was never
faster than not splitting it. Parallelism therefore divides by region, and
sequences handed in directly are staged to a temporary file so that the
workers read them rather than receive them.

# Session information

```{r session-info, eval=TRUE}
sessionInfo()
```
