---
title: "Runtime and Memory for Genome-Scale Tm Profiling: an hg38 Benchmark Guide"
author: "Junhui Li, Lihua Julie Zhu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Runtime and Memory for Genome-Scale Tm Profiling: an hg38 Benchmark Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,   # requires BSgenome.Hsapiens.UCSC.hg38 and long runtimes
  message = FALSE,
  warning = FALSE
)
```

# Introduction

This vignette answers two practical questions that come up as soon as Tm
profiling is scaled from a bacterial genome to the human genome (hg38):

1. For a **single chromosome** (chr1), how long does the pipeline take, how
   much memory does it use, and why does supplying a multi-worker
   `BPPARAM` to `tm_calculate()` **not** make it faster?
2. For the **whole genome**, how should the work be parallelized, what are
   sensible parameter settings, and what runtime and memory should be
   budgeted?

All timings below were measured on a 6-core (12-thread) Intel MacBook Pro
with 16 GB RAM, R 4.4, macOS, with the package's compiled (Rcpp)
nearest-neighbor core. Absolute numbers will differ on other machines; the
*structure* of the results — which parts scale, which parts do not — will
not. None of the chunks are evaluated when the vignette is built; run them
interactively to reproduce the measurements on your own hardware.

```{r libraries}
library(TmCalculator)
library(BiocParallel)

pkg <- "BSgenome.Hsapiens.UCSC.hg38"
suppressPackageStartupMessages(library(pkg, character.only = TRUE))
genome <- get(pkg, envir = asNamespace(pkg))
```

# Measuring time and memory

Three tools cover almost every need. Wall-clock time comes from
`system.time()` (read the `elapsed` column; for parallel runs `user` and
`system` only reflect the manager process). The peak of R-managed memory
over a code block comes from the `gc(reset = TRUE)` idiom. The true
resident size of the current process — including C++ allocations that R's
GC does not see — comes from `ps::ps_memory_info()`.

```{r measuring}
gc(reset = TRUE)                      # reset the "max used" high-water mark
## ... run the code to be measured ...
gc()                                  # read peak from the "max used" column

ps::ps_memory_info()[["rss"]] / 1e9   # current process resident set, GB
```

(Earlier releases exported a `gc()` of their own, for the GC *content* of a
sequence, which masked R's garbage collector once the package was attached.
That function is now `gc_content()`, so no `base::` prefix is needed.)

For `SnowParam()` workers the manager cannot see worker memory at all;
record it inside the worker function and return it with the result (this
pattern is used in the whole-genome section below).

# A ten-second warm-up: chr21

Before committing to chr1, verify your setup on chr21 — at 46.7 Mb it is
the smallest hg38 autosome and the whole pipeline finishes in about ten
seconds. It is also a deliberately instructive choice: chr21 *begins*
with a ~5 Mb assembly gap of `N` bases, so it exercises the N-trimming
machinery that most chromosomes only touch lightly.

```{r chr21-warmup}
chr_len21 <- GenomeInfoDb::seqlengths(genome)[["chr21"]]

t21 <- system.time({
  bins21 <- make_genomiccoord(bsgenome = pkg, chromosomes = "chr21",
                              window = 200L, slide = 200L,
                              start = 1, end = chr_len21, strand = "+")
  gr21   <- to_genomic_ranges_fast(list(pkg_name = pkg, seq = bins21),
                                   method = "preload_chr")
  tm21   <- tm_calculate(gr21, method = "tm_nn",
                         nn_table = "DNA_NN_SantaLucia_2004", Na = 50)
})
t21["elapsed"]        # ~10-15 s cold, ~6 s warm, on the test machine
head(tm21$gr)
```

Watch the `make_genomiccoord()` messages: N-trimming reports an effective
range starting at position 5,010,001 — the leading 5 Mb of `N`s is
detected in a single in-memory scan and excluded before any windows are
built, leaving 208,449 windows instead of 233,549. The first run in a
session pays one-time costs (reading the genome from disk, loading the
thermodynamic tables); subsequent runs are roughly twice as fast. If this
warm-up runs cleanly, your BSgenome installation, compiled core and
memory headroom are all in working order, and the chr1 numbers below
should reproduce within noise.

One benchmarking caveat that bites easily: `devtools::load_all()`
compiles the C++ core at `-O0` (a debug build, ~6x slower) and leaves a
mixed session state even after reinstalling. Always benchmark in a fresh
session with a plain `library(TmCalculator)`; confirm with
`"devtools_shims" %in% search()` returning `FALSE`.

# Part 1: a single chromosome (chr1)

## The pipeline and its serial cost

chr1 is 248,956,422 bp. Non-overlapping 200 bp windows after N-trimming
give 1,244,682 windows.

```{r chr1-serial}
chr_len <- GenomeInfoDb::seqlengths(genome)[["chr1"]]

t_coord <- system.time({
  bins <- make_genomiccoord(bsgenome = pkg, chromosomes = "chr1",
                            window = 200L, slide = 200L,
                            start = 1, end = chr_len, strand = "+")
})

t_extract <- system.time({
  gr_batch <- to_genomic_ranges_fast(list(pkg_name = pkg, seq = bins),
                                     method = "preload_chr")
})

base::gc(reset = TRUE)
t_tm <- system.time({
  tm_chr1 <- tm_calculate(gr_batch, method = "tm_nn",
                          nn_table = "DNA_NN_SantaLucia_2004", Na = 50)
})
base::gc()   # "max used" = peak R memory during the Tm step

rbind(t_coord, t_extract, t_tm)[, "elapsed"]
## t_coord t_extract      t_tm
##     5.5      15.4      30.8     (freshly booted, otherwise idle machine)
```

Measured on the test machine (`R --vanilla`, freshly booted, no other
applications):

| step | function | elapsed | notes |
|---|---|---|---|
| window coordinates | `make_genomiccoord()` | ~5.5 s | N-end scan loads the 249 Mb chromosome once |
| sequence extraction | `to_genomic_ranges_fast(method = "preload_chr")` | ~15 s | `extractAt()` itself is ~4 s; the rest is complement generation, the GC column and `GRanges` assembly |
| Tm calculation | `tm_calculate(method = "tm_nn")` | ~31 s | compiled core, 1.24M windows |
| **total** | | **~52 s** | |

On a machine with normal background load the same steps run 20-35%
slower (we measured 7.3 / 19.3 / 41.5 s in an ordinary desktop session);
comparisons within one section below were always measured in a single
session.

Memory: `base::gc()` reported a peak of ~1.9 GB of R-managed heap during
the Tm step, and a dedicated worker process running the whole chr1
pipeline reaches **~3.5-4 GB resident** (sequences, complements, a
character copy handed to the compiled core, `GRanges` metadata, plus
loaded packages). A useful planning figure: **budget ~4 GB per
chr1-sized pipeline**, tapering to well under 2 GB for small
chromosomes.

Two extraction details matter at this scale. Pass the input as
`list(pkg_name = ..., seq = ...)` so the genome package is loaded once,
and use `method = "preload_chr"` for dense tiling of few chromosomes
(`getSeq()`-based `"vectorized"` is better for windows scattered across
many chromosomes).

## Why `BPPARAM` does not accelerate a single chromosome

`tm_calculate()` accepts a `BPPARAM` and it does exactly what it
advertises — yet on chr1 it does not help:

```{r chr1-parallel}
system.time({
  tm_serial <- tm_calculate(gr_batch, method = "tm_nn", Na = 50)
})
## elapsed ~ 30 s

system.time({
  tm_snow <- tm_calculate(gr_batch, method = "tm_nn", Na = 50,
                          BPPARAM = SnowParam(workers = 5))
})
## elapsed ~ 55 s on the idle machine (33-80 s across repeated sessions)
## -- NEVER faster than serial
```

The serial time is stable across sessions; the parallel time is not,
because its added costs (spawning workers, serializing half a gigabyte
of sequence, loading packages in each worker) are highly sensitive to
machine state. The decomposition explains why even its best case cannot
win. Of the ~30 s serial runtime, only the per-window Tm loop (~9 s with
the compiled core) is divided among workers. The rest is inherently
serial or is *new* cost created by parallelization:

| component | ~time | scales with workers? |
|---|---|---|
| serial preprocessing (N filtering, coercion, result assembly) | ~21 s | no — runs once in the manager |
| shipping ~0.5 GB of sequence to workers and results back | ~5 s | no — serialization is serial |
| worker startup + package loading | ~3+ s | fixed cost, *added* by parallelism |
| compiled Tm loop | ~9 s / n workers | **yes** |

This is Amdahl's law with an unusually small parallel fraction (~30%):
even with infinitely many workers the floor is ~29 s — essentially the
serial runtime — and on a loaded machine the added costs balloon well
past it. With the pure-R core of previous releases the loop
dominated (minutes for an input of this size) and the same
`SnowParam(5)` call gave a 2.8x speedup; compiling the loop moved the
bottleneck into the serial part and made within-call parallelism
unprofitable. **With the compiled core, the serial default
(`SerialParam()`) is the fastest way to run inputs up to roughly a
million windows.**

A stronger warning applies to `MulticoreParam()` on large inputs. Forked
workers share the manager's memory copy-on-write, but R's garbage
collector *writes* to every object header it marks, so each worker's GC
forces the kernel to physically duplicate inherited pages. On a
chr1-scale input (measured with the pure-R core, where the effect is
easiest to see), `MulticoreParam(5)` ran *slower than serial* while
burning ~18 CPU-minutes of `system` time on page duplication. If you
parallelize at all, prefer `SnowParam()`.

# Part 2: the whole genome

## Parallelize at the chromosome level

The serial preprocessing that defeats within-call parallelism is itself
embarrassingly parallel *across chromosomes*: window building, extraction,
filtering and assembly for chr2 are independent of chr1. So for
genome-scale work, give each worker one whole chromosome and run the
entire pipeline — coordinates, extraction, `tm_calculate()` with the
serial default — inside the worker:

```{r genome-parallel}
chrs <- paste0("chr", c(1:22, "X", "Y"))
## Better: sort largest-first for load balance -- see "Choosing the
## worker count" below for the one-liner.
n_workers <- 5   # see "Choosing the worker count" below

runtime <- system.time({
  res_list <- bplapply(chrs, function(chr, pkg) {
    ## SnowParam workers are fresh R processes: load packages HERE,
    ## not in the manager session.
    suppressPackageStartupMessages(library(TmCalculator))
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))

    genome  <- get(pkg, envir = asNamespace(pkg))
    chr_len <- GenomeInfoDb::seqlengths(genome)[[chr]]

    bins <- make_genomiccoord(bsgenome = pkg, chromosomes = chr,
                              window = 200L, slide = 200L,
                              start = 1, end = chr_len, strand = "+",
                              verbose = FALSE)
    gr  <- to_genomic_ranges_fast(list(pkg_name = pkg, seq = bins),
                                  method = "preload_chr")
    out <- tm_calculate(gr, method = "tm_nn",
                        nn_table = "DNA_NN_SantaLucia_2004",
                        Na = 50)$gr        # serial inside the worker

    ## Drop sequence columns before returning: Tm/GC are what we keep,
    ## and this cuts per-chromosome serialization from ~500 MB to a few MB.
    out$sequence   <- NULL
    out$complement <- NULL

    ## Record this worker's peak resident memory (GB) for budgeting.
    attr(out, "worker_rss_gb") <- ps::ps_memory_info()[["rss"]] / 1e9
    out
  }, pkg = pkg, BPPARAM = SnowParam(workers = n_workers))

  tm_genome <- unlist(GenomicRanges::GRangesList(res_list))
})

runtime
sapply(res_list, attr, "worker_rss_gb")   # per-worker memory check
length(tm_genome)                         # 14,687,330 windows
summary(tm_genome$Tm)
## Min. 47.2  1st Qu. 68.9  Median 71.5  Mean 72.0  3rd Qu. 74.8  Max. 101.1
```

Unlike within-call `BPPARAM`, this scheme parallelizes *everything*:
window building, extraction, filtering and assembly, not just the Tm
loop. Per-task communication is a chromosome name in and a slim
`GRanges` out, so Amdahl's law has little serial residue to bite on.

The chromosome is not the only unit of work available, though, and the
choice of unit turns out to matter more than the worker count. The next
section defines three schemes and the one after it measures all three
over the same worker range.

## Three ways to divide the work

The three differ in *how many* tasks there are and in *when* a task is
assigned to a worker.

**Static dispatch** is `SnowParam(workers = n)` left alone. BiocParallel
pre-splits the task list into one contiguous chunk per worker before
anything runs, so a worker that draws a light chunk finishes early and
then sits idle for the rest of the run.

**Dynamic dispatch** is the same task list with `tasks = length(X)`.
The manager hands out one element at a time, so a worker that finishes
immediately takes the next. Sorting the chromosomes largest first makes
this effective: the long tasks are placed while there is still short
work left to fill the ragged end of the run.

**Segment dispatch** keeps dynamic assignment and changes the unit.
Instead of 24 whole chromosomes the genome is cut into 50 Mb intervals,
giving 73 tasks, and the longest single task falls from 55 s to 12 s.
No schedule can finish sooner than its longest task, so this lowers the
floor that the other two schemes are pressed against.

```{r strategies}
sl   <- GenomeInfoDb::seqlengths(genome)[paste0("chr", c(1:22, "X", "Y"))]
chrs <- names(sort(sl, decreasing = TRUE))          # largest first

tasks_chrom <- lapply(chrs, function(ch)
  list(chr = ch, start = 1L, end = as.integer(sl[[ch]])))

## Segment boundaries must be multiples of `slide`, or the window grid
## shifts between segments and the result stops matching a whole-
## chromosome run. 50 Mb = 250,000 x 200 bp.
seg_size  <- 50e6
tasks_seg <- unlist(lapply(names(sl), function(ch) {
  st <- seq(1, sl[[ch]], by = seg_size)
  lapply(st, function(s)
    list(chr = ch, start = as.integer(s),
         end = as.integer(min(s + seg_size - 1, sl[[ch]]))))
}), recursive = FALSE)

length(tasks_chrom)   # 24
length(tasks_seg)     # 73

n <- 5
BPPARAM_static  <- SnowParam(workers = n)
BPPARAM_dynamic <- SnowParam(workers = n, tasks = length(tasks_chrom))
BPPARAM_segment <- SnowParam(workers = n, tasks = length(tasks_seg))
```

One correctness detail: a segment task must pass `trim_N = "none"` to
`make_genomiccoord()`, while a whole-chromosome task should trim. Trimming
inside a segment would move the window grid relative to the chromosome and
the two schemes would no longer compute the same windows. Interior `N`
windows are dropped afterwards in both cases, so the totals agree exactly.

## The measured sweep

All 24 chromosomes, 14,687,330 windows, 200 bp non-overlapping, freshly
booted, `R --vanilla`. Wall time excludes worker start-up, which is
reported separately because workers pay it concurrently and it therefore
contributes its longest instance rather than the sum. Speedup is the
one-worker wall time of the same strategy divided by the wall time of
the configuration.

```{r sweep-data, eval=TRUE}
# The numbers below are the measured sweep. They are held in a data frame
# rather than typed into a markdown table so that the table and the figure
# that follow cannot drift apart: both are rendered from this object.
sweep <- data.frame(
  strategy = rep(c("static", "dynamic", "segment"), each = 6),
  workers  = rep(1:6, 3),
  tasks    = rep(c(24L, 24L, 73L), each = 6),
  wall_s   = c(590.5, 431.9, 328.0, 270.0, 251.0, 244.6,
               589.1, 318.2, 237.4, 209.8, 236.4, 262.0,
               578.5, 320.2, 227.9, 193.9, 180.0, 194.0),
  startup_s = c(0.0, 7.1, 7.9, 8.1, 9.0, 9.7,
                0.0, 7.4, 7.2, 8.2, 8.7, 9.9,
                0.0, 6.9, 7.8, 8.3, 9.0, 9.8),
  work_s   = c(590.5, 613.9, 638.2, 656.5, 730.0, 906.3,
               589.1, 620.7, 683.9, 795.8, 1093.9, 1481.9,
               578.5, 629.1, 665.3, 736.9, 842.6, 1085.9),
  longest_task_s = c(55.2, 57.1, 60.8, 61.9, 70.7, 92.4,
                     53.4, 59.4, 63.6, 69.7, 88.2, 127.5,
                     11.5, 13.8, 15.6, 15.7, 17.5, 22.6),
  rss_gb   = c(4.75, 4.06, 4.06, 3.85, 3.70, 3.35,
               4.74, 3.86, 3.38, 2.90, 2.67, 2.15,
               4.87, 2.58, 2.58, 2.34, 1.86, 1.77),
  stringsAsFactors = FALSE)

# Derived rather than transcribed. Speedup is the one-worker wall clock of
# the SAME strategy over the configuration's wall clock: segmenting changes
# how much total work there is, so a ratio formed against a configuration's
# own summed task times would credit a strategy for doing more work and the
# three could not be compared. Efficiency follows from that speedup, not
# from the task times.
serial <- with(sweep, tapply(wall_s[workers == 1], strategy[workers == 1], I))
sweep$speedup    <- serial[sweep$strategy] / sweep$wall_s
sweep$efficiency <- sweep$speedup / sweep$workers
serial_work <- with(sweep, tapply(work_s[workers == 1], strategy[workers == 1], I))
sweep$work_ratio <- sweep$work_s / serial_work[sweep$strategy]
sweep$strategy <- factor(sweep$strategy, levels = c("static", "dynamic", "segment"))
sweep <- sweep[order(sweep$strategy, sweep$workers), ]
```

All 24 chromosomes, 14,687,330 windows, 200 bp non-overlapping, freshly
booted, `R --vanilla`.

```{r sweep-table, eval=TRUE}
tab <- data.frame(
  Strategy   = ifelse(duplicated(sweep$strategy), "", as.character(sweep$strategy)),
  Workers    = sweep$workers,
  Tasks      = sweep$tasks,
  `Wall (s)` = round(sweep$wall_s, 1),
  `Start-up (s)` = round(sweep$startup_s, 1),
  `Total task time (s)` = round(sweep$work_s, 1),
  `Work ratio` = round(sweep$work_ratio, 2),
  `Longest task (s)` = round(sweep$longest_task_s, 1),
  Speedup    = round(sweep$speedup, 2),
  Efficiency = round(sweep$efficiency, 2),
  `Peak RSS per worker (GB)` = round(sweep$rss_gb, 2),
  check.names = FALSE, stringsAsFactors = FALSE)
knitr::kable(tab, row.names = FALSE,
             caption = "Task-partitioning strategy and worker count.")
```

```{r sweep-figure, eval=TRUE, fig.width=10.5, fig.height=3.9, fig.cap="Parallel performance under three task-partitioning strategies. (A) Speedup against worker count, relative to the one-worker run of the same strategy; the dashed line marks linear speedup. (B) Total task time, summed over all tasks and measured inside the workers; values above the dashed serial reference indicate that a configuration performed more work than the serial run rather than dividing the same work among more processes. (C) Peak resident set size of the heaviest worker."}
# Three panels, one argument. Panel A shows that every strategy turns over
# before the cores run out, which invites the usual explanation of a ragged
# schedule. Panel B rules that out: the total task time grows with the worker
# count, so the later configurations are doing more work, not dividing a
# fixed amount badly. Panel C gives the reason and the remedy together, since
# the strategy that holds its memory down is the one whose work grows least.
pal  <- c(static = "#1B5E9C", dynamic = "#C0392B", segment = "#5C6B73")
pchs <- c(static = 16,        dynamic = 17,        segment = 15)
lv   <- levels(sweep$strategy)

op <- par(mfrow = c(1, 3), mar = c(4.4, 4.5, 2.2, 0.8), las = 1,
          cex = 0.8, mgp = c(2.8, 0.7, 0))

series <- function(col, ylab, ref = NULL, ref_lab = NULL, ymax = NULL) {
  plot(NA, xlim = range(sweep$workers),
       ylim = c(0, if (is.null(ymax)) max(sweep[[col]]) * 1.05 else ymax),
       bty = "n", xaxt = "n", xlab = "Workers", ylab = ylab)
  axis(1, at = sort(unique(sweep$workers)))
  if (!is.null(ref)) {
    abline(h = ref, lty = 2, col = "grey55")
    text(max(sweep$workers), ref, ref_lab, adj = c(1.1, -0.5),
         cex = 0.75, col = "grey40")
  }
  for (s in lv) {
    d <- sweep[sweep$strategy == s, ]
    lines(d$workers, d[[col]], col = pal[s], lwd = 2)
    points(d$workers, d[[col]], col = pal[s], pch = pchs[s])
  }
}

series("speedup", "Speedup", ymax = max(sweep$workers) * 1.02)
abline(a = 0, b = 1, lty = 2, col = "grey55")
text(max(sweep$workers), max(sweep$workers), "linear",
     adj = c(1.1, -0.4), cex = 0.75, col = "grey40")
for (s in lv) {                      # redraw over the reference line
  d <- sweep[sweep$strategy == s, ]
  lines(d$workers, d$speedup, col = pal[s], lwd = 2)
  points(d$workers, d$speedup, col = pal[s], pch = pchs[s])
}
mtext("A", side = 3, adj = 0, font = 2, line = 0.7, cex = 1.05)
legend("topleft", bty = "n", legend = lv, col = pal[lv], pch = pchs[lv],
       lwd = 2, seg.len = 1.3, cex = 0.85)

series("work_s", "Total task time (s)",
       ref = median(sweep$work_s[sweep$workers == 1]), ref_lab = "serial")
mtext("B", side = 3, adj = 0, font = 2, line = 0.7, cex = 1.05)

series("rss_gb", "Peak resident set size per worker (GB)")
mtext("C", side = 3, adj = 0, font = 2, line = 0.7, cex = 1.05)

par(op)
```

Three readings.

**Dispatch order matters at every worker count.** Static and dynamic run
the identical 24 tasks; the only difference is that static commits them
to workers up front. At four workers that costs 60 s, a fifth of the run.

**Segmenting is worth more than dispatch order.** Segment dispatch is
faster than dynamic at every worker count from three upwards, and its
best time, 180 s, is 30 s better than dynamic's best and 65 s better
than static's. The mechanism is in the "longest task" column: 55 s of
chr1 cannot be divided by any scheduler, and at five workers that single
task is nearly a third of the whole run.

**Segmenting halves the memory.** A worker holding a 50 Mb segment needs
about 1.9 GB against 3.7 GB for one holding chr1, and peak memory is what
decides how many workers a machine can run at all. On this 16 GB laptop
segment dispatch is both the fastest scheme and the one that leaves the
most headroom, which is an unusually convenient combination.

**All three turn over well before the cores run out.** Six physical cores
are available, yet static peaks at 6, dynamic at 4 and segment at 5, and
none reaches half of linear speedup. The next section explains why, and
the reason is not the one that first suggests itself.

## Why the speedup saturates

The obvious explanation for a curve that flattens is a ragged schedule:
one long task holding up the end while other workers idle. The "total
task time" column rules that out. It is the sum of the per-task times,
measured inside the workers and excluding their package loading, so it is
the amount of work the run actually performed. If the schedule were
merely uneven, that column would stay flat and only the wall clock would
suffer.

It does not stay flat. From one worker to six it grows by 53% under
static dispatch, 88% under segment dispatch and 152% under dynamic
dispatch. The later configurations are not dividing a fixed amount of
work badly, they are doing more of it: the same chr1 task that takes
53 s alone takes 127 s with five siblings competing for memory bandwidth
and last-level cache. Each worker is individually slower, and past a
certain point adding one slows the others by more than it contributes.

This also explains the ordering. Dynamic dispatch inflates worst because
its tasks are the largest resident objects, so the workers contend most
for cache; segmenting reduces the resident footprint per worker and
inflates less. The effect is a property of the machine rather than of
the software, so a cluster node with more memory channels will show the
turnover later, or not at all. Rerun the sweep on your own hardware
rather than transplanting these worker counts.

The `work ratio` column is the one to watch when you do. As long as it
stays near 1.0, adding workers is buying real parallelism; once it passes
about 1.3 the run is mostly paying for its own contention.

## Where the extra work lands

The summary table reports only the longest task, which hides which task
that was and whether the inflation is spread evenly. The benchmark also
writes a per-task file, `*_tasks.csv`, with one row per task and
repetition: chromosome, interval, window count, time, and the worker's
resident memory. Recovering the per-chromosome picture from it is a few
lines:

```{r per-chrom}
tk <- utils::read.csv("bench_parallel_strategy_tasks.csv")
tk <- tk[tk$strategy == "dynamic", ]        # one task per chromosome

## Median over repetitions, one column per worker count.
w <- stats::reshape(
  stats::aggregate(secs ~ chr + n_workers, data = tk, FUN = stats::median),
  idvar = "chr", timevar = "n_workers", direction = "wide")
names(w) <- sub("^secs\\.", "w", names(w))

w$inflation <- w$w6 / w$w1                  # 6 workers vs serial
w <- w[order(-w$w1), ]
w
```

Read it two ways. Down the `w1` column is the serial cost of each
chromosome, which is what sets the floor no scheduler can go below: the
largest single value is the hard limit on any chromosome-level schedule,
and it is the number segmenting exists to reduce. Across a row is how
much slower that same task became as workers were added, with `inflation`
summarising it. A ratio near 1.0 means the task was indifferent to its
neighbours; a large ratio means it was competing with them for memory
bandwidth and cache.

The distinction matters for choosing a strategy. If inflation were
uniform across chromosomes it would be a fixed tax and segmenting would
not help with it. If instead the large chromosomes inflate most, then the
same property that makes them stragglers also makes them the main source
of contention, and cutting them into segments addresses both at once.

## Segmenting one chromosome

Segment dispatch also answers the question Part 1 left open: how *do*
you parallelize a single chromosome? Ship coordinates, not sequences.
Each worker receives a `(start, end)` pair, a few dozen bytes, and runs
the entire pipeline on its own segment. Every row of the Part 1 cost
table becomes parallel except worker startup, because the preprocessing
happens inside the workers and the half-gigabyte serialization never
occurs.

```{r segment-parallel}
chr_len <- GenomeInfoDb::seqlengths(genome)[["chr1"]]

## Segment length must be a multiple of `slide` so the window grid stays
## aligned across segment boundaries (50 Mb = 250,000 x 200).
seg_starts <- seq(1, chr_len, by = 50e6)
seg <- data.frame(start = seg_starts,
                  end   = pmin(seg_starts + 50e6 - 1, chr_len))

t_seg <- system.time({
  res_seg <- bplapply(seq_len(nrow(seg)), function(i, seg, pkg) {
    suppressPackageStartupMessages(library(TmCalculator))
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))
    bins <- make_genomiccoord(bsgenome = pkg, chromosomes = "chr1",
                              window = 200L, slide = 200L,
                              start = seg$start[i], end = seg$end[i],
                              strand = "+", trim_N = "none",
                              verbose = FALSE)
    gr <- to_genomic_ranges_fast(list(pkg_name = pkg, seq = bins),
                                 method = "preload_chr")
    out <- tm_calculate(gr, method = "tm_nn",
                        nn_table = "DNA_NN_SantaLucia_2004", Na = 50)$gr
    out$sequence <- NULL; out$complement <- NULL
    out
  }, seg = seg, pkg = pkg, BPPARAM = SnowParam(workers = 5))

  tm_chr1_seg <- sort(unlist(GenomicRanges::GRangesList(res_seg)))
})

t_seg["elapsed"]      # 38 s measured, vs ~52 s for the serial FULL pipeline
length(tm_chr1_seg)   # 1,152,300
```

Mind the comparison scope: 38 s covers the *entire* pipeline (windows,
extraction, Tm), so the serial reference is the ~52 s pipeline total from
Part 1 — a **1.35x** speedup — not the 30 s Tm-only figure. Modest, but
it is the only single-chromosome parallelization that wins at all
(within-call `SnowParam` *lost* to serial above), and the result is
verifiably identical: 1,152,300 windows equals the serial run's
1,244,682 generated minus its 92,382 N-gap windows, exactly. That
equivalence is by construction — `trim_N = "none"` keeps the window grid
identical across segments (the "Skipped ... regions" warning from
`tm_calculate()` is expected; hg38 chr1 carries an ~18 Mb interior
centromeric gap), and the 50 Mb segment length is a multiple of `slide`.

The gain is capped well below worker-count because each worker pays
fixed costs the serial run pays once: loading packages and reading the
same 249 Mb chromosome to extract its segment, with five workers
contending for the same file. Per-worker memory stays moderate, about
1 to 2 GB, since only a 50 Mb segment is windowed at a time.

## Choosing the strategy and the worker count

Take segment dispatch. It was fastest at every worker count from three
upwards, it halves the per-worker memory, and the memory saving is what
lets a machine run more workers in the first place. Chromosome-level
dispatch is worth keeping only if the downstream analysis wants whole
chromosomes back as single objects.

For the worker count, two constraints, take the minimum:

```{r workers}
mem_gb    <- 16                                     # your machine
cores     <- parallel::detectCores(logical = FALSE) # physical cores
per_worker_gb <- 2                                  # 50 Mb segment task
n_workers <- min(cores - 1L, floor((mem_gb - 4) / per_worker_gb))
n_workers
## 16 GB / 6 cores  -> 5 workers, which is what the sweep found fastest
## 32 GB / 8 cores  -> 7 workers
## 64 GB / 10 cores -> 9 workers
```

Use 2 GB per worker for segment tasks and 4 GB for whole-chromosome
tasks; those are the measured peaks, not estimates. Do not count
hyperthreads. The compiled core is CPU-bound and logical cores add little
beyond scheduler overhead, and the sweep above shows the run is already
contending for cache at five real cores.

Treat the memory bound as hard. When workers do collide the system
swaps, and a swapping parallel run is slower than serial: we measured
19 GB of swap and a wedged machine in exactly this state, aggravated by
leftover workers from interrupted runs.

Two smaller settings that the sweep shows are not optional. Sort the
tasks largest-first, because the scheduler deals them out in order and a
long task placed last has nothing left to overlap with. And pass
`tasks = length(X)` to `SnowParam()`; leaving it at the default is what
makes a run "static", and that cost 60 s at four workers here for no
benefit at all.

```{r sort-chrs}
sl   <- GenomeInfoDb::seqlengths(genome)
chrs <- names(sort(sl[paste0("chr", c(1:22, "X", "Y"))], decreasing = TRUE))
```

## Reproducing the sweep

The whole measurement is scripted, including the per-task instrumentation
that separates worker start-up from compute and records each worker's
peak resident memory:

```sh
Rscript inst/scripts/bench_parallel_strategy.R \
  --workers 1,2,3,4,5,6 --reps 3 --outfile bench_parallel_strategy.csv

Rscript inst/scripts/make_table6.R  --csv bench_parallel_strategy.csv
Rscript inst/scripts/make_figure5.R --csv bench_parallel_strategy.csv
```

Include one worker in the sweep. It is the baseline every speedup is
divided by, and BiocParallel runs a single worker inside the manager
rather than spawning one, so that point measures the serial cost with
nothing to serialise. Without it the only available ratios are against
each strategy's own summed task times, which are not comparable across
strategies because segmenting changes how much total work there is.

For a cluster, `inst/scripts/bench_parallel_cluster.R` adds the three
things a shared machine needs: it reads the allocated core count from the
scheduler rather than from `detectCores()`, which on a shared node reports
the whole machine and oversubscribes badly; it stages the genome package
on node-local disk, because BSgenome reads the `.2bit` at random offsets
and several workers hitting a shared filesystem measure storage latency
instead of software; and it sweeps physical cores only.

## Housekeeping

Four operational notes. First, benchmark in a session that did *not*
restore a saved workspace (`[Workspace loaded from ~/.RData]` at startup
means gigabytes of old objects are already resident); disable
"Restore .RData at startup" in RStudio or start `R --vanilla`. Second,
`SnowParam()` workers load the *installed* package: after
`devtools::load_all()` development changes, run `devtools::install()`
before benchmarking, and note that `load_all()` compiles C++ at `-O0` —
time only the installed build. The same applies to
`devtools::build_vignettes()`, which evaluates vignette code against a
`-O0` debug build (timings come out ~6x slow); set
`options(pkg.build_extra_flags = FALSE)` to disable the debug flags, or
rely on `R CMD build` / `devtools::check()`, which install with normal
optimization before running vignettes. Changing the flag option does
*not* trigger recompilation — stale `-O0` object files in `src/` are
reused as long as sources are unchanged — so run `pkgbuild::clean_dll()`
after changing it.
Third, if a parallel run errors or is interrupted, workers can linger
and hold gigabytes; check for orphaned `R` processes (Activity Monitor on
macOS, `ps aux | grep R` elsewhere) before rerunning. Fourth, worker
`message()` output is buffered until each task finishes — a silent
console does not mean a hung run; watch the worker processes' CPU
instead.

# Summary

For hg38 with the compiled nearest-neighbor core, run single chromosomes
serially. Leave `BPPARAM` at its default: within-call parallelism cannot
beat Amdahl's law when the parallel fraction is about 30%.

For the whole genome, parallelize outside the call, over 50 Mb segments
rather than whole chromosomes, with `SnowParam(workers = n, tasks =
length(X))` and the tasks sorted largest-first. On a 6-core, 16 GB
laptop that runs 14.7 million windows in 180 s at five workers, against
578 s serially, a speedup of 3.21 at 1.9 GB per worker. The same run
organised by chromosome takes 210 s at best and needs 2.9 GB per worker.

Set the worker count from `min(physical cores - 1, (memory - 4) / 2)`,
then verify rather than trust it. Every strategy here turned over before
the cores ran out, and the reason was not scheduling: the total task time
grew with the worker count, by 53% to 152% from one worker to six, as
workers competed for memory bandwidth and cache. That is a property of
the machine, so the right worker count on your hardware is the one the
sweep finds, not the one copied from this table.

# Session info

```{r sessioninfo, eval=TRUE}
sessionInfo()
```
