---
title: "Genome-Scale Tm Profiling with tm_profile()"
author: "Junhui Li, Lihua Julie Zhu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genome-Scale Tm Profiling with tm_profile()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,   # the hg38 chunks need BSgenome.Hsapiens.UCSC.hg38 and minutes
  message = FALSE,
  warning = FALSE,
  fig.retina = 1,
  dpi        = 72
)
```

`tm_profile()` builds a melting temperature profile from a sequence source
and a set of regions. It tiles the regions into windows, fetches each
window's sequence, computes a Tm, and returns one `GRanges`. On a mammalian
genome it spreads the work across processes, which is the difference
between ten minutes and three.

This vignette covers the two things that matter in practice: what you can
ask it to profile, and how many workers to give it. The computation chunks
are not evaluated when the vignette is built, since they need
`BSgenome.Hsapiens.UCSC.hg38` and take minutes; the timing table and figure
are built from measurements that ship with the package.

```{r libraries}
library(TmCalculator)
library(BiocParallel)

hg38 <- "BSgenome.Hsapiens.UCSC.hg38"

tm <- tm_profile(hg38, window = 200, slide = 200,
                 method = "tm_nn", nn_table = "DNA_NN_SantaLucia_2004",
                 Na = 50, BPPARAM = SnowParam(workers = 5))
length(tm)
## [1] 14687330
```

With no `regions` argument that covers the standard chromosomes, chr1 to
chr22, chrX and chrY. Leave `BPPARAM` out and the same call runs in one
process, with no dependency on **BiocParallel** at all: how the genome is
divided changes the runtime, not the windows.

# What you can profile

The first argument is where the sequence comes from, given by *name* so
that each worker can open it itself: an installed BSgenome package, or the
path to a FASTA file.

```{r sources}
tm_profile(hg38)                       # a BSgenome package
tm_profile("contigs.fa.gz")            # a FASTA assembly, gzip is fine
tm_profile(oligos)                     # sequences already in R
```

The third form is staged as a temporary FASTA under `tmpdir` and removed on
exit, so that the workers read the sequences rather than receive them. That
is worth doing for a large set, because it moves window construction and
result assembly into the worker too; for a handful of sequences the staging
and the worker start-up cost more than the calculation, and `tm_calculate()`
is the right call. Set `tmpdir` on a cluster, where the default `tempdir()`
is often a small partition.

`regions` says what to take from it, in whichever form is at hand.

```{r regions}
## Chromosomes or records, by name or by number. On a BSgenome the chr
## prefix is added or removed to match the genome, so the same code works
## on UCSC and on Ensembl; FASTA record names are matched exactly.
tm_profile(hg38, regions = 1:22)
tm_profile(hg38, regions = c("chr1", "chr2"))
tm_profile("contigs.fa.gz", regions = c("contig_7", "contig_9"))

## Coordinate intervals. Commas and scientific notation are accepted, so a
## number pasted out of a genome browser works as it stands.
tm_profile(hg38, regions = c("chr1:1-10e6", "chrX:5,000,000-6,000,000"))

## A mixture, when some chromosomes are wanted whole and others in part.
tm_profile(hg38, regions = c(1:20, "chr21:1-10e6", "X", "Y"))

## A GRanges, when the regions come from an annotation.
prom <- promoters(genes(TxDb.Hsapiens.UCSC.hg38.knownGene),
                  upstream = 1000, downstream = 500)
tm_profile(hg38, regions = prom, window = 50, slide = 25)

## window = NULL gives one window per region, which is what short records
## call for: a FASTA of array probes, primers or synthetic oligos returns
## one Tm per record. It is the default when sequences are passed directly.
tm_profile("probes.fa", window = NULL, method = "tm_nn", Na = 50)
tm_profile(oligos, method = "tm_nn", Na = 50, BPPARAM = SnowParam(5))
```

Two behaviours are worth knowing. A region naming a whole chromosome has
its leading and trailing assembly gaps trimmed, since a telomeric run of N
carries no windows; a region given by coordinate is tiled from the start
you asked for. And overlapping regions produce windows that appear twice,
so `tm_profile()` warns rather than quietly double-counting them.

Array manifests are not read directly, since their layout is vendor and
version specific. Take the probe sequences out with the package that
already understands the manifest, `illuminaio`, `minfi` or `sesame` for
Infinium arrays, then pass them as a FASTA file or straight to
`tm_calculate()`.

# Dividing the work

A task is the unit a worker owns from start to finish: it opens the source,
builds its own windows, fetches its own sequence and computes its own Tm.
Only a name and a coordinate pair cross between processes, so no sequence
is ever serialized. `unit` decides how the tasks are cut.

```{r unit}
tm_profile(hg38, unit = "segment", segment_size = 50e6,   # default: 73 tasks
           BPPARAM = SnowParam(workers = 5))
tm_profile(hg38, unit = "region",                         # 24 tasks
           BPPARAM = SnowParam(workers = 5))
```

Segments are the better default. Chromosome 1 is an indivisible task of
249 Mb, so with one task per chromosome the run cannot finish before
chromosome 1 does, however many workers are available; and a worker holding
a 50 Mb segment needs less memory than one holding a whole chromosome.
Segment boundaries fall on multiples of `slide`, so the window grid, and
therefore the result, is the same either way.

Prefer `SnowParam()` to `MulticoreParam()`. Forked workers share the
manager's memory copy-on-write, but R's garbage collector writes to every
object header it marks, so each worker's collection forces the kernel to
duplicate the inherited pages; on a genome-scale input that has been
measured running slower than a single process.

# How many workers

Memory is usually the binding constraint rather than cores, because each
worker holds the sequences of its own task. Below is the sweep that ships
with the package: hg38 at 200 bp non-overlapping windows, 14,687,330
windows, one to six workers, three repetitions per configuration, on a
six-core 16 GB laptop and on a cluster compute node with six slots and a
48 GB memory limit.

```{r sweep-data, eval=TRUE}
# Read the shipped sweeps rather than transcribing numbers, so the table
# cannot drift from the measurements it describes. The two files do not
# carry the same columns, the cluster sweep also recording the host and CPU
# model it landed on, so both are cut to the four columns used here before
# they are stacked.
cols <- c("strategy", "n_workers", "wall_s", "max_rss_gb")
read_sweep <- function(file, env) {
  d <- utils::read.csv(system.file("extdata", file, package = "TmCalculator"),
                       stringsAsFactors = FALSE)
  d <- d[d$strategy == "segment", cols]      # the default unit of tm_profile()
  d$env <- env
  d
}
raw <- rbind(read_sweep("bench_parallel_strategy.csv", "Laptop"),
             read_sweep("bench_parallel_cluster.csv",  "Compute node"))

sweep <- do.call(rbind, lapply(
  split(raw, list(raw$env, raw$n_workers), drop = TRUE),
  function(d) data.frame(
    Environment = d$env[1], Workers = d$n_workers[1],
    wall_s = median(d$wall_s), lo = min(d$wall_s), hi = max(d$wall_s),
    rss_gb = median(d$max_rss_gb), stringsAsFactors = FALSE)))

serial <- stats::setNames(sweep$wall_s[sweep$Workers == 1L],
                          sweep$Environment[sweep$Workers == 1L])
sweep$Speedup <- serial[sweep$Environment] / sweep$wall_s
sweep <- sweep[order(sweep$Environment != "Laptop", sweep$Workers), ]
```

```{r sweep-table, eval=TRUE}
knitr::kable(
  data.frame(
    Environment = sweep$Environment,
    Workers     = sweep$Workers,
    `Wall time (s)` = sprintf("%.1f [%.1f-%.1f]", sweep$wall_s, sweep$lo, sweep$hi),
    Speedup     = sprintf("%.2f", sweep$Speedup),
    `Peak RSS per worker (GB)` = sprintf("%.2f", sweep$rss_gb),
    check.names = FALSE),
  row.names = FALSE,
  caption = paste("Median of three repetitions, observed range in brackets.",
                  "Wall time includes worker start-up."))
```

```{r sweep-figure, eval=TRUE, fig.width=8, fig.height=3.6, fig.cap="Wall-clock time and peak memory per worker against worker count, on a six-core 16 GB laptop (solid, filled) and a compute node with a 48 GB limit (dashed, open). Points are medians of three repetitions; bars give the observed range."}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.4, 2.2, 0.8), mgp = c(2.6, 0.7, 0))
for (what in c("wall", "rss")) {
  ys <- if (what == "wall") sweep$wall_s else sweep$rss_gb
  plot(range(sweep$Workers), range(0, ys * 1.05), type = "n",
       xlab = "Workers", xaxt = "n", adj = 0,
       ylab = if (what == "wall") "Wall clock (s)" else "Peak resident size per worker (GB)",
       main = if (what == "wall") "A" else "B")
  axis(1, at = sort(unique(sweep$Workers)))
  for (e in unique(sweep$Environment)) {
    d <- sweep[sweep$Environment == e, ]
    y <- if (what == "wall") d$wall_s else d$rss_gb
    solid <- e == "Laptop"
    if (what == "wall")
      arrows(d$Workers, d$lo, d$Workers, d$hi, angle = 90, code = 3,
             length = 0.03, col = "grey40")
    lines(d$Workers, y, lty = if (solid) 1 else 2, col = "grey20")
    points(d$Workers, y, pch = if (solid) 19 else 1, col = "grey20")
  }
  if (what == "wall")
    legend("topright", c("Laptop, 16 GB", "Compute node, 48 GB"),
           lty = c(1, 2), pch = c(19, 1), bty = "n", cex = 0.85, col = "grey20")
}
par(op)
```

Two things in that table decide a worker count.

The laptop is fastest at five workers, 189 s against 578.5 s in one
process, and a sixth worker makes the run *slower*. The node, with three
times the memory, improves all the way to six workers and 127 s. The
turnover is memory pressure rather than anything in the software, which is
why the optimum belongs to the machine and not to the package.

Memory per worker falls as the work is spread more widely, from 4.87 GB in
the one-worker run to 1.86 GB at five workers, because no worker holds more
than one 50 Mb task at a time. That is what makes more workers affordable
at all.

A rule that works on both machines:

> workers = min(physical cores - 1, (available GB - 4) / 2)

Then confirm it on your own machine with a short sweep over one chromosome
rather than the whole genome:

```{r sweep-your-machine}
for (n in 1:6) {
  t <- system.time(
    tm_profile(hg38, regions = "chr21", window = 200, slide = 200,
               method = "tm_nn", segment_size = 10e6,
               BPPARAM = SnowParam(workers = n), verbose = FALSE))
  cat(sprintf("%d workers: %.1f s\n", n, t[["elapsed"]]))
}
```

The sweeps above were produced by `bench_parallel_strategy.R` and
`bench_parallel_cluster.R` in `system.file("scripts", package =
"TmCalculator")`, and their output is the two CSV files read here, so
re-running one on your own hardware means replacing a file.

# What comes back, and what to use instead

A `GRanges` with `Tm` and `GC` metadata columns in genomic order, ready for
`integrate_granges()`, `compare_groups()` and the plotting functions, and
for anything else that takes genomic intervals. The sequence and complement
columns are dropped by default, since they run to roughly 500 MB per large
chromosome; pass `keep_sequence = TRUE` if you need them.

`tm_profile()` starts from coordinates in a source. When the sequences are
already in hand, as a character vector, a FASTA file or a `GRanges` with a
`sequence` column, use `tm_calculate()`:

```{r tm-calculate}
tm_calculate(c("ACGTGCTAGCTAGCTAGC", "GGCCATATATGCGC"),
             method = "tm_nn", Na = 50)
```

Do not try to parallelise a single `tm_calculate()` call. With the compiled
core the per-window loop is a minority of a call's cost; window
construction, sequence retrieval and result assembly run once in the
calling process, and sending the sequences to workers costs more than the
loop it divides. Measured on chromosome 1, splitting one call across five
workers was never faster than the serial call. Parallelism belongs at the
level of regions, which is what `tm_profile()` does.

# Session information

```{r session-info, eval=TRUE}
sessionInfo()
```
