---
title: "TmCalculator: A Comprehensive Tool for Melting Temperature Calculations"
author: "Junhui Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TmCalculator: A Comprehensive Tool for Melting Temperature Calculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)

library(TmCalculator)
library(BSgenome.Hsapiens.UCSC.hg38)
library(GenomicRanges)
library(IRanges)
```

# Introduction

The TmCalculator package provides comprehensive options for calculating melting temperatures (Tm) of DNA/RNA sequences. This vignette demonstrates the package's functionality, including:

- Basic sequence input and Tm calculation with the Wallace rule (`tm_wallace` via `tm_calculate()`)
- Genomic coordinate input and Tm calculation with the nearest-neighbor method (`tm_nn`)
- FASTA file input and Tm calculation
- Complementary sequence input (mismatches and dangling ends)
- Genome-wide visualization with `plot_karyotype_genome()`, `plot_tm_genome_tracks()`, `plot_tm_heatmap()`, and `plot_tm_linear()`
- Interactive plotly views via `plot_tm_*_interactive()` helpers
- An overview of the optional Shiny application

## Installation

```{r install, eval = FALSE}
suppressMessages({
  library(BiocManager)
  BiocManager::install(c("Biostrings", "BSgenome", "BSgenome.Hsapiens.UCSC.hg38"))
})
install.packages("TmCalculator")
pak::pkg_install("Gviz")
pak::pkg_install("JunhuiLi1017/TmCalculator@dev")
```

# Basic Usage

## Simple Sequence Input with Wallace rule method

```{r basic_usage, message=FALSE}
seqs <- c("ATGCGCGAAAT", "GCTAGCTAGCT")
names(seqs) <- c("chr1:1-11:+:seq1", "chr2:1-11:+:seq2")
result <- tm_calculate(seqs, method = "tm_wallace")
result$gr
```

## Genomic Coordinate Input with nearest neighbor method

Coordinates should follow `"chr:start-end:strand:genome_pkg:seq_name"`.

```{r genomic_input, message=FALSE}
coords <- c(
  "chr1:1898000-1898050:+:BSgenome.Hsapiens.UCSC.hg38:seq1",
  "chr2:2563000-2563050:-:BSgenome.Hsapiens.UCSC.hg38:seq2"
)
result <- tm_calculate(coords, method = "tm_nn")
result$gr
```

### Nearest-neighbor options

```{r genomic_input_options, message=FALSE}
result$options
```

## FASTA File Processing

```{r fasta_processing, message=FALSE}
fasta_file <- system.file("extdata", "example1.fasta", package = "TmCalculator")
result_fa <- tm_calculate(fasta_file, method = "tm_nn")
result_fa$gr
```

## Complementary Sequence with a mismatch

GCTAGCCGA[C]AATGGGCAGATAGTAGAAA  
CGATCGGCT[A]TTACCCGTCTATCATCTTT

```{r complement_mismatch, message=FALSE}
seqs <- c("GCTAGCCGACAATGGGCAGATAGTAGAAA")
comp_seqs <- c("CGATCGGCTATTACCCGTCTATCATCTTT")
result <- tm_calculate(
  input_seq = seqs,
  complement_seq = comp_seqs,
  method = "tm_nn"
)
result$gr
```

## Complementary Sequence with a dangling end and a mismatch

GCTAGCCGA[C]AATGGGCAGATAGTAGAAA  
 GATCGGCT[A]TTACCCGTCTATCATCTTT

```{r complement_dangling_end, message=FALSE}
seqs <- c("GCTAGCCGACAATGGGCAGATAGTAGAAA")
comp_seqs <- c("GATCGGCTATTACCCGTCTATCATCTTT")
result <- tm_calculate(
  input_seq = seqs,
  complement_seq = comp_seqs,
  method = "tm_nn",
  shift = -1
)
result$gr
```

## Genome tiling with `make_genomiccoord()` and `to_genomic_ranges_fast()`

For large tiling jobs, generate coordinate strings with `make_genomiccoord()` and convert them with the fast backend:

```{r genome_tiling, eval = FALSE}
coords <- make_genomiccoord(
  bsgenome    = BSgenome.Hsapiens.UCSC.hg38,
  chromosomes = "chr1",
  window      = 200L,
  slide       = 50L,
  trim_N      = "ends",
  verbose     = FALSE
)
gr <- to_genomic_ranges_fast(
  list(pkg_name = "BSgenome.Hsapiens.UCSC.hg38", seq = head(coords, 20))
)
tm_tiled <- tm_calculate(gr, method = "tm_nn")
```

# Visualization of Tm values

Visualization functions expect a `GRanges` object with a numeric `Tm` metadata column (and optionally `GC` from `tm_calculate()`).

```{r tm_viz_data, message=FALSE}
set.seed(123)

chr1_starts <- sort(sample(1:249250621, 100))
chr1_ends   <- chr1_starts + sample(50:200, 100, replace = TRUE)
chr2_starts <- sort(sample(1:243199373, 50))
chr2_ends   <- chr2_starts + sample(50:200, 50, replace = TRUE)

tm_results <- GRanges(
  seqnames = Rle(c(rep("chr1", 100), rep("chr2", 50))),
  ranges = IRanges(
    start = c(chr1_starts, chr2_starts),
    end   = c(chr1_ends, chr2_ends)
  ),
  strand = Rle(sample(c("+", "-"), 150, replace = TRUE)),
  Tm = c(runif(100, 60, 80), runif(50, 60, 80)),
  GC = c(runif(100, 30, 70), runif(50, 30, 70))
)
genome(seqinfo(tm_results)) <- "hg38"

gr_tm <- GRanges(
  seqnames = c("chr1", "chr2", "chr1", "chr2", "chr1"),
  ranges = IRanges(
    start = c(100, 200, 300, 400, 150),
    end   = c(150, 250, 350, 450, 200)
  ),
  Tm = c(65.5, 68.2, 70.1, 63.8, 72.0)
)
```

## Karyotype multi-track view (`plot_karyotype_genome`)

`plot_karyotype_genome()` uses **karyoploteR** with a flexible `track_list` (same idea as `plot_circos_genome()`).

```{r plot_karyotype, eval=FALSE}
track_list <- list(
  list(type = "points", data = tm_results, value_col = "Tm",
       col = "#e41a1c", name = "Tm (deg C)"),
  list(type = "bars", data = tm_results, value_col = "GC",
       col = "#377eb8", name = "GC (%)")
)

plot_karyotype_genome(
  genome_assembly = "hg38",
  track_list      = track_list,
  chromosomes     = c("chr1", "chr2"),
  title_name      = "Tm and GC - hg38"
)

plot_karyotype_genome(
  genome_assembly = "hg38",
  track_list = list(
    list(type = "line", data = tm_results[seqnames(tm_results) == "chr1"],
         value_col = "Tm", col = "#984ea3", name = "Tm (deg C)")
  ),
  chromosomes = "chr1",
  zoom        = "chr1:50000000-150000000",
  title_name  = "Tm - chr1 zoom"
)
```

## Genome-browser tracks (`plot_tm_genome_tracks`)

`plot_tm_genome_tracks()` uses **Gviz** for a genome-browser layout, or **ggplot2** in `x_axis = "regions"` mode for sparse probes.

```{r plot_genome_tracks, eval=FALSE}
plot_tm_genome_tracks(
  tm_results,
  chromosome_to_plot = "chr1",
  genome_assembly    = "hg38",
  track_type         = "gradient",
  zoom               = "chr1:10000000-80000000"
)

p_regions <- plot_tm_genome_tracks(
  tm_results,
  chromosome_to_plot = "chr1",
  genome_assembly    = "hg38",
  x_axis             = "regions",
  track_type         = "points",
  color_by           = "Tm"
)
print(p_regions)
```

## Heatmap views (`plot_tm_heatmap`)

```{r plot_heatmap, eval=FALSE}
plot_tm_heatmap(gr_tm, genome_assembly = "hg19", plot_type = "karyogram")
plot_tm_heatmap(gr_tm, genome_assembly = "hg19", plot_type = "faceted")
plot_tm_heatmap(gr_tm, genome_assembly = "hg19", x_axis = "regions")
```

## Linear region plot (`plot_tm_linear`)

```{r plot_linear, fig.height=5}
plot_tm_linear(
  gr_tm,
  color_by     = "Tm",
  facet_by_chr = FALSE,
  add_line     = TRUE
)
```

## Interactive plots

```{r plot_interactive, eval=FALSE}
plot_tm_heatmap_interactive(gr_tm, genome_assembly = "hg19", plot_type = "karyogram")
plot_tm_genome_tracks_interactive(
  tm_results,
  chromosome_to_plot = "chr1",
  genome_assembly    = "hg38",
  x_axis             = "regions"
)
plot_tm_karyotype_interactive(
  tm_results,
  genome_assembly = "hg38",
  chromosomes     = c("chr1", "chr2")
)
```

## Launch the Shiny application

```{r launch_shiny, eval = FALSE}
TmCalculatorShiny::TmCalculator_shiny()
```

# Conclusion

The TmCalculator package provides tools for calculating and analyzing melting temperatures of DNA/RNA sequences. Use `plot_karyotype_genome()` and `plot_tm_genome_tracks()` for genome-context views, `plot_tm_heatmap()` for interactive overviews, and `plot_tm_linear()` for compact region-by-region comparisons.

For more information, see the package documentation and function examples.
