---
title: "Window-Size Sensitivity of Genome-Wide Tm Profiles"
author: "Junhui Li, Lihua Julie Zhu"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Window-Size Sensitivity of Genome-Wide Tm Profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 6
)
```

# Introduction

Any window-based genomic profile depends on the window size, which acts as a
smoothing bandwidth: too large and local signal is averaged away, too small
and single-window noise dominates. This vignette quantifies that dependence
for **TmCalculator** by recomputing the *Escherichia coli* K-12 MG1655
Tm/GC profile at 50, 100, 200 and 500 bp and asking two separate questions.

1. Does window size change the thermodynamic landscape itself?
2. Does it change a biological conclusion drawn from that landscape?

The companion vignette
`vignette("genome_wide_tm_ecoli", package = "TmCalculator")` develops the
full case study at the 200 bp resolution used by Hasenauer *et al.* (2025),
including the MutL-associated region (MutL-AR) comparison that is re-tested
here at every window size.

# Setup

We reproduce the minimum needed from the main case study: the genome, the
MutL-AR peak coordinates, and the reference labels used for plotting.

```{r packages}
library(TmCalculator)
library(GenomicRanges)
library(IRanges)
library(GenomeInfoDb)
```

The *E. coli* genome is supplied by the pre-forged
`BSgenome.Ecoli.NCBI.ASM584v2` package. The chunk below installs it if it is
not already available, exactly as in the main case-study vignette; see that
vignette for how the package is forged from the NCBI assembly accession.

```{r setup-ecoli-bsgenome, message=FALSE, warning=FALSE}
ecoli_pkg  <- "BSgenome.Ecoli.NCBI.ASM584v2"
genome_obj <- "Ecoli"   # BSgenomeObjname in DESCRIPTION; not the package name

.ecoli_genome_ready <- function() {
  if (!requireNamespace(ecoli_pkg, quietly = TRUE)) return(FALSE)
  exists(genome_obj, envir = asNamespace(ecoli_pkg), inherits = FALSE)
}

if (!.ecoli_genome_ready()) {
  if (!requireNamespace("remotes", quietly = TRUE)) {
    utils::install.packages("remotes", repos = "https://cloud.r-project.org")
  }
  remotes::install_github(
    "JunhuiLi1017/BSgenome.Ecoli.NCBI.ASM584v2",
    upgrade = "never",
    quiet = TRUE
  )
}

if (!.ecoli_genome_ready()) {
  stop(
    "Could not load genome object '", genome_obj, "' from package '",
    ecoli_pkg, "'.\n",
    "See vignette(\"genome_wide_tm_ecoli\") for how to forge it locally.",
    call. = FALSE
  )
}
```

```{r load-genome}
suppressPackageStartupMessages(library(ecoli_pkg, character.only = TRUE))
genome      <- base::get(genome_obj, envir = asNamespace(ecoli_pkg))
genome_name <- ecoli_pkg
chr_name    <- "U00096.3"
chr_length  <- GenomeInfoDb::seqlengths(genome)[[chr_name]]

data(ecoli_rep_hotspots)

## MutL-AR peaks, as in the main case study
mutH_peaks <- GRanges(
  seqnames = ecoli_rep_hotspots$all_peaks_IP_mutH$chr,
  ranges   = IRanges(start = ecoli_rep_hotspots$all_peaks_IP_mutH$start,
                     end   = ecoli_rep_hotspots$all_peaks_IP_mutH$end)
)
seqlevels(mutH_peaks) <- chr_name
mutH_peaks$peak_id <- paste0("mutH_", seq_along(mutH_peaks))

## Reference labels: replication origin (ori) and terminus (dif)
label <- data.frame(
  seqnames = genome_name,
  start    = c(3925804, 1590777),
  end      = c(3925804, 1590777),
  label    = c("ori", "dif")
)
```

---

# Sensitivity analysis

For each window size we retile the chromosome, recompute Tm and GC, and
re-run the MutL-AR comparison, keeping every other setting fixed. Nothing
downstream needs to change: the annotation layers keep their own native
resolutions (1 kb bins for the microsatellite, cruciform and GATC density
tracks; variable-width intervals for MutL-AR peaks and ssDNA regions), and
both `plot_genome_track()` and the overlap-based statistics work across
mixed resolutions.

```{r window-sensitivity}
window_sizes <- c(50L, 100L, 200L, 500L)

sens <- lapply(window_sizes, function(w) {
  bins_w <- make_genomiccoord(
    bsgenome = genome_name, chromosomes = chr_name,
    window = w, slide = w, start = 1, end = chr_length,
    strand = "+", verbose = FALSE
  )
  gr_w <- to_genomic_ranges_fast(list(pkg_name = genome_name, seq = bins_w))
  tm_w <- tm_calculate(gr_w, method = "tm_nn",
                       nn_table = "DNA_NN_Breslauer_1986", Na = 50)$gr

  ann <- integrate_granges(gr_tm = tm_w, gr_features = mutH_peaks,
                           strategy = "overlap", feature_cols = "peak_id",
                           keep_unmatched = TRUE)
  ann$in_mutH <- ifelse(is.na(ann$peak_id), "non_peak", "peak")
  cg <- compare_groups(gr = ann, target = c("Tm", "GC"),
                       method = "wilcoxon", group = "in_mutH",
                       alternative = "greater", posthoc = FALSE)
  list(gr = tm_w, ann = ann, test = cg)
})
names(sens) <- paste0("w", window_sizes)
```

### Distributional stability

```{r sensitivity-distribution}
dist_tbl <- do.call(rbind, lapply(seq_along(window_sizes), function(i) {
  g <- sens[[i]]$gr
  data.frame(
    window_bp = window_sizes[i],
    n_windows = length(g),
    Tm_mean   = mean(g$Tm, na.rm = TRUE),
    Tm_sd     = stats::sd(g$Tm, na.rm = TRUE),
    Tm_IQR    = stats::IQR(g$Tm, na.rm = TRUE),
    GC_mean   = mean(g$GC, na.rm = TRUE)
  )
}))
knitr::kable(dist_tbl, digits = 3,
             caption = "Tm/GC distribution by window size.")
```

Two systematic effects are visible, both expected. First, mean Tm rises
with window length (`r sprintf("%.1f", dist_tbl$Tm_mean[1])` to
`r sprintf("%.1f", dist_tbl$Tm_mean[nrow(dist_tbl)])` degrees C from
`r dist_tbl$window_bp[1]` to `r dist_tbl$window_bp[nrow(dist_tbl)]` bp):
duplex stability increases with length in the nearest-neighbor model, so
absolute Tm values are only comparable *within* one window size. Second,
shorter windows widen the distribution (SD
`r sprintf("%.2f", dist_tbl$Tm_sd[1])` at `r dist_tbl$window_bp[1]` bp vs
`r sprintf("%.2f", dist_tbl$Tm_sd[nrow(dist_tbl)])` at
`r dist_tbl$window_bp[nrow(dist_tbl)]` bp) -- window size acts as a
smoothing bandwidth. The mean GC is invariant
(`r sprintf("%.3f", mean(dist_tbl$GC_mean))` throughout), confirming that
the composition landscape itself does not depend on the tiling.

### Cross-scale agreement

We aggregate each finer profile onto the 500 bp grid (mean Tm per bin)
and correlate it with the directly computed 500 bp profile:

```{r sensitivity-correlation}
gr500  <- sens[["w500"]]$gr
bin500 <- (start(gr500) - 1L) %/% 500L

cor_tbl <- vapply(c("w50", "w100", "w200"), function(k) {
  g   <- sens[[k]]$gr
  agg <- tapply(g$Tm, (start(g) - 1L) %/% 500L, mean, na.rm = TRUE)
  stats::cor(agg[match(bin500, as.integer(names(agg)))], gr500$Tm,
             use = "complete.obs")
}, numeric(1))
round(cor_tbl, 3)
```

Aggregated onto a common 500 bp grid, the profiles are nearly
interchangeable (r =
`r paste(sprintf("%.3f", cor_tbl), collapse = ", ")` for 50, 100 and
200 bp respectively). Window size rescales and smooths the profile but
preserves the spatial landscape.

### Robustness of the MutL-AR association

Finally, the vignette's biological conclusion -- that Tm and GC differ
between MutL-AR peak windows and the genomic background -- is re-tested
at every window size:

```{r sensitivity-tests}
## Per-window-size test statistics
test_results <- do.call(rbind, lapply(seq_along(window_sizes), function(i) {
  data.frame(window_bp = window_sizes[i], sens[[i]]$test$results)
}))
test_results

## Per-window-size group summaries, plus the median Tm effect size
test_summary <- do.call(rbind, lapply(seq_along(window_sizes), function(i) {
  ann <- sens[[i]]$ann
  med <- tapply(ann$Tm, ann$in_mutH, stats::median, na.rm = TRUE)
  data.frame(window_bp   = window_sizes[i],
             n_peak      = sum(ann$in_mutH == "peak"),
             Tm_med_diff = unname(med["peak"] - med["non_peak"]),
             sens[[i]]$test$summary)
}))
test_summary
```

### Multi-scale tracks alongside the multi-omics layers

Because every track keeps its native resolution, the four Tm/GC profiles
can be drawn as concentric layers of one integrated figure, together
with the annotation data:

```{r sensitivity-tracks, fig.width=8, fig.height=8, fig.cap="Multi-scale integration. Concentric rings from outside in: MutL-AR peaks (ideogram), GC content at 50/100/200/500 bp (blues), Tm at 50/100/200/500 bp (reds), and microsatellite density (green). Yellow bands mark MutL-AR peaks."}
## darkest = finest window (50 bp), lightest = coarsest (500 bp)
scale_cols_gc <- c("#1A5276", "#2E86C1", "#5DADE2", "#AED6F1")
scale_cols_tm <- c("#7B241C", "#CB4335", "#EC7063", "#F5B7B1")

sens_dfs <- lapply(sens, function(x) as.data.frame(x$gr[, c("Tm", "GC")]))

tracks_scale <- c(
  list(list(type = "rect", data = ecoli_rep_hotspots$all_peaks_IP_mutH,
            col = "#2C3E50", bg.col = "grey", name = "MutL-AR",
            legend_font_col = "#2C3E50", ideogram = TRUE, height = 0.5)),
  lapply(seq_along(window_sizes), function(i)
    list(type = "line", data = sens_dfs[[i]], value_col = "GC",
         name = paste0("GC ", window_sizes[i], " bp"),
         col = scale_cols_gc[i], legend_font_col = scale_cols_gc[i])),
  lapply(seq_along(window_sizes), function(i)
    list(type = "line", data = sens_dfs[[i]], value_col = "Tm",
         name = paste0("Tm ", window_sizes[i], " bp"),
         col = scale_cols_tm[i], legend_font_col = scale_cols_tm[i])),
  list(list(type = "line", data = ecoli_rep_hotspots$bins_rep,
            value_col = "count", name = "Microsatellites", col = "#2ECC71",
            legend_font_col = "#2ECC71"),
       list(type = "highlight",
            data = ecoli_rep_hotspots$all_peaks_IP_mutH,
            col = "#F1C40F", alpha = 0.18))
)

plot_genome_track(
  genome_name = genome_name,
  genome_size = chr_length,
  track_list  = tracks_scale,
  circular    = TRUE,
  label       = label
)
```

At whole-genome scale the four bandwidths look nearly identical -- the
landscape is invariant. The difference window size makes is local
smoothing, best seen in a zoomed view:

```{r sensitivity-zoom, fig.width=10, fig.height=7, fig.cap="Zoomed 1.5-1.7 Mb region. The 50 bp profile resolves single-window fluctuations that the 500 bp profile averages away; all four trace the same underlying landscape."}
plot_genome_track(
  genome_name = genome_name,
  genome_size = chr_length,
  track_list  = tracks_scale,
  zoom        = "U00096.3:1500000-1700000"
)
```

```{r sensitivity-effect-sizes, echo=FALSE}
eff <- unique(test_summary[, c("window_bp", "Tm_med_diff")])
gc_rows <- test_summary[test_summary$target == "GC", ]
```

The association is reproduced at every scale. MutL-AR windows are lower
in median Tm than the background at all four window sizes
(`r paste(sprintf("%+.2f", eff$Tm_med_diff), collapse = ", ")` degrees C
at `r paste(eff$window_bp, collapse = ", ")` bp) and lower in GC
(`r sprintf("%.3f", mean(gc_rows$mean[gc_rows$group == "peak"]))` in peaks
vs `r sprintf("%.3f", mean(gc_rows$mean[gc_rows$group == "non_peak"]))`
in the background), and every Wilcoxon test is significant
(largest p = `r format(max(test_results$p.value), digits = 2, scientific = TRUE)`).
As expected, p values shrink with the number of windows while the effect
size stays essentially constant, so the biological conclusion does not
depend on the window choice. The primary analysis uses 200 bp to match
Hasenauer *et al.* (2025), whose Methods compute GC content and melting
temperature in 200 bp bins (using TmCalculator) and bin the ChIP-seq
coverage at the same 200 bp resolution.

---

# Session Information

```{r session-info}
sessionInfo()
```
